
# A Single Heatmap {#a-single-heatmap}

A single heatmap is the most used approach for visualizing the data. Although
"the shining point" of the **ComplexHeatmap** package is that it can visualize a
list of heatmaps in parallel, however, as the basic unit of the heatmap list, it is
still very important to have the single heatmap well configured.

First let's generate a random matrix where there are three groups in the columns
and three groups in the rows:

```{r data}
set.seed(123)
nr1 = 4; nr2 = 8; nr3 = 6; nr = nr1 + nr2 + nr3
nc1 = 6; nc2 = 8; nc3 = 10; nc = nc1 + nc2 + nc3
mat = cbind(rbind(matrix(rnorm(nr1*nc1, mean = 1,   sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc1, mean = 0,   sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc1, mean = 0,   sd = 0.5), nr = nr3)),
    rbind(matrix(rnorm(nr1*nc2, mean = 0,   sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc2, mean = 1,   sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc2, mean = 0,   sd = 0.5), nr = nr3)),
    rbind(matrix(rnorm(nr1*nc3, mean = 0.5, sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc3, mean = 0.5, sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc3, mean = 1,   sd = 0.5), nr = nr3))
   )
mat = mat[sample(nr, nr), sample(nc, nc)] # random shuffle rows and columns
rownames(mat) = paste0("row", seq_len(nr))
colnames(mat) = paste0("column", seq_len(nc))
```

Following command contains the minimal argument for the `Heatmap()` function
which visualizes the matrix as a heatmap with default settings. Very
similar as other heatmap tools, it draws the dendrograms, the row/column names
and the heatmap legend. The default color schema is "blue-white-red" which is
mapped to the minimal-mean-maximal values in the matrix. The title for the
legend is assigned with an internal index number.

```{r default, fig.width = 5.2}
Heatmap(mat)
```

The title for the legend is taken from the "name" of the heatmap by default.
Each heatmap has a name which is like a unique identifier for the heatmap and
it is important when you have a list of heatmaps. In later chapters, you will
find the heatmap name is used for setting the "main heatmap" and is used for
decoration of heatmaps. If the name is not assigned, an internal name is
assigned to the heatmap in a form of `matrix_%d`. In following examples in
this chapter, we give the name `mat` to the heatmap (for which you will see
the change of legend in the next plot).

If you put `Heatmap()` inside a function or a `for`/`if`/`while` chunk, you
won't see the heatmap after executing `Heatmap()`. In this case, you need to
use `draw()` function explicitly as follows. We will explain this point in
more detail in Section \@ref(plot-the-heatmap).

```{r, eval = FALSE}
for(...) {
    ht = Heatmap(mat)
    draw(ht)
}
```

## Colors {#colors}

For heatmap visualization, colors are the major representation of the data
matrix. In most cases, the heatmap visualizes a matrix with continuous numeric
values. In this case, users should provide a color mapping function. A color
mapping function should accept a vector of values and return a vector of
corresponding colors. **Users should always use `circlize::colorRamp2()`
function to generate the color mapping function** in `Heatmap()`. The
two arguments for `colorRamp2()` is a vector of break values and a vector of
corresponding colors. `colorRamp2()` linearly interpolates colors in every
interval through LAB color space. Also using `colorRamp2()` helps to generate
a legend with proper tick marks.

In following example, values between -2 and 2 are linearly interpolated to get
corresponding colors, values larger than 2 are all mapped to red and values
less than -2 are all mapped to green.

```{r, fig.width = 5.2}
library(circlize)
col_fun = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
col_fun(seq(-3, 3))
Heatmap(mat, name = "mat", col = col_fun)
```

As you can see, the color mapping function exactly maps negative values to
green and positive values to red, even when the distribution of negative
values and positive values are not centric to zero. Also this color mapping
function is **not affected by outliers**. In following plot, the clustering is
heavily affected by the outlier (see the dendrogram) but not the color mapping.

```{r, fig.width = 5.2}
mat2 = mat
mat2[1, 1] = 100000
Heatmap(mat2, name = "mat", col = col_fun, 
    column_title = "a matrix with outliers")
```

More importantly, `colorRamp2()` makes colors in multiple heatmaps comparible
if they are set with a same color mapping function. In following three
heatmaps, a same color always corresponds to a same value.

```{r, eval = FALSE}
Heatmap(mat, name = "mat", col = col_fun, column_title = "mat")
Heatmap(mat/4, name = "mat", col = col_fun, column_title = "mat/4")
Heatmap(abs(mat), name = "mat", col = col_fun, column_title = "abs(mat)")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 10/3}
pushViewport(viewport(x = 0, width = 1/3, just = "left"))
draw(Heatmap(mat, name = "mat", col = col_fun, column_title = "mat"), newpage = FALSE)
popViewport()
pushViewport(viewport(x = 1/3, width = 1/3, just = "left"))
draw(Heatmap(mat/4, name = "mat", col = col_fun, column_title = "mat/4"), newpage = FALSE)
popViewport()
pushViewport(viewport(x = 2/3, width = 1/3, just = "left"))
draw(Heatmap(abs(mat), name = "mat", col = col_fun, column_title = "abs(mat)"), newpage = FALSE)
popViewport()
```

If the matrix is continuous, you can also simply provide a vector of colors
and colors will be linearly interpolated. But remember this method is not
robust to outliers because the mapping starts from the minimal value in the
matrix and ends with the maximal value. Following color mapping setting is
identical to `colorRamp2(seq(min(mat), max(mat), length = 10), rev(rainbow(10)))`.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", col = rev(rainbow(10)), 
    column_title = "set a color vector for a continuous matrix")
```

If the matrix contains discrete values (either numeric or character), colors
should be specified as a named vector to make it possible for the mapping from
discrete values to colors. If there is no name for the color, the order of
colors corresponds to the order of `unique(mat)`. Note now the legend is
generated from the color mapping vector.

Following sets colors for a discrete numeric matrix (you don't need to convert
it to a character matrix).

```{r, fig.width = 5.2}
discrete_mat = matrix(sample(1:4, 100, replace = TRUE), 10, 10)
colors = structure(1:4, names = c("1", "2", "3", "4")) # black, red, green, blue
Heatmap(discrete_mat, name = "mat", col = colors,
    column_title = "a discrete numeric matrix")
```

Or a character matrix:

```{r discrete_character_matrix, fig.width = 5.2}
discrete_mat = matrix(sample(letters[1:4], 100, replace = TRUE), 10, 10)
colors = structure(1:4, names = letters[1:4])
Heatmap(discrete_mat, name = "mat", col = colors,
    column_title = "a discrete character matrix")
```

As you see in the two examples above, for the numeric matrix (no matter the
color is continuous mapping or discrete mapping), by default clustering is
applied on both dimensions while for character matrix, clustering is turned
off (but you can still cluster a character matrix if you provide a proper
distance metric for two character vectors, see example in Section
\@ref(distance-methods)).

``NA`` is allowed in the matrix. You can control the color of `NA` by `na_col`
argument (by default it is grey for `NA`). **The matrix that contains `NA` can
be clustered by `Heatmap()`.**

Note the `NA` value is not presented in the legend.

```{r, fig.width = 5.2}
mat_with_na = mat
na_index = sample(c(TRUE, FALSE), nrow(mat)*ncol(mat), replace = TRUE, prob = c(1, 9))
mat_with_na[na_index] = NA
Heatmap(mat_with_na, name = "mat", na_col = "black",
    column_title = "a matrix with NA values")
```

Color space is important for interpolating colors. By default, colors are
linearly interpolated in [LAB color
space](https://en.wikipedia.org/wiki/Lab_color_space), but you can select the
color space in `colorRamp2()` function. Compare following two plots. Can you
see the difference?

```{r, eval = FALSE}
f1 = colorRamp2(seq(min(mat), max(mat), length = 3), c("blue", "#EEEEEE", "red"))
f2 = colorRamp2(seq(min(mat), max(mat), length = 3), c("blue", "#EEEEEE", "red"), space = "RGB")
Heatmap(mat, name = "mat1", col = f1, column_title = "LAB color space")
Heatmap(mat, name = "mat2", col = f2, column_title = "RGB color space")
```

```{r, fig.width = 10, echo = FALSE}
f1 = colorRamp2(seq(min(mat), max(mat), length = 3), c("blue", "#EEEEEE", "red"))
f2 = colorRamp2(seq(min(mat), max(mat), length = 3), c("blue", "#EEEEEE", "red"), 
    space = "RGB")
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(Heatmap(mat, name = "mat1", col = f1, column_title = "LAB color space"), newpage = FALSE)
upViewport()
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(Heatmap(mat, name = "mat2", col = f2, column_title = "RGB color space"), newpage = FALSE)
upViewport()
```

In following plots, corresponding values change evenly on the folded lines,
you can see how colors change under different color spaces (top plots:
green-black-red, bottom plots: blue-white-red. The plot is made by
[**HilbertCurve**package](https://bioconductor.org/packages/release/bioc/html/HilbertCurve.html)).

```{r, fig.width = 14, fig.height = 14/5, echo = FALSE, message = FALSE}
space = c("RGB", "LAB", "XYZ", "sRGB", "LUV")
pushViewport(viewport(layout = grid.layout(nr = 1, nc = length(space))))
for(i in seq_along(space)) {
    pushViewport(viewport(layout.pos.row = 1, layout.pos.col = i))
    hc = HilbertCurve::HilbertCurve(1, 100, level = 4, newpage = FALSE, title = space[i])
    ir = IRanges::IRanges(start = 1:99, end = 2:100)
    f = colorRamp2(c(-1, 0, 1), c("green", "black", "red"), space = space[i])
    col = f(seq(-1, 1, length = 100))
    HilbertCurve::hc_points(hc, ir, np = 3, gp = gpar(col = col, fill = col))
    upViewport()
}
upViewport()
grid.newpage()
pushViewport(viewport(layout = grid.layout(nr = 1, nc = length(space))))
for(i in seq_along(space)) {
    pushViewport(viewport(layout.pos.row = 1, layout.pos.col = i))
    hc = HilbertCurve::HilbertCurve(1, 100, level = 4, newpage = FALSE, title = space[i])
    ir = IRanges::IRanges(start = 1:99, end = 2:100)
    f = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"), space = space[i])
    col = f(seq(-1, 1, length = 100))
    HilbertCurve::hc_points(hc, ir, np = 3, gp = gpar(col = col, fill = col))
    upViewport()
}
upViewport()
```

Last but not the least, colors for the heatmap borders can be set by the
`border`/`border_gp` and `rect_gp` arguments. `border`/`border_gp` controls
the global border of the heatmap body and `rect_gp` controls the border of the
grids/cells in the heatmap.

The value of `border` can be logical (`TRUE` corresponds to `black`) or a
character of color (e.g. `red`). The use for `border` argument is only for historical
reason, here you can also set `border_gp` argument which should be a `gpar` object.

`rect_gp` is a `gpar` object which means you can only set it by
`grid::gpar()`. Since the filled color is already controlled by the heatmap
color mapping, you can only set the `col` parameter in `gpar()` to control the
border of the heatmap grids.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", border_gp = gpar(col = "black", lty = 2),
    column_title = "set heatmap borders")
Heatmap(mat, name = "mat", rect_gp = gpar(col = "white", lwd = 2),
    column_title = "set cell borders")
```

If `col` is not set, the default color mapping by `Heatmap()` is designed with
trying to be as convinient and meaningful as possible. Following are the rules
for the default color mapping (by `ComplexHeatmap:::default_col()`):

- If the values are characters, the colors are generated by
  `circlize::rand_color()`.
- If the values are from the heatmap annotation and are numeric, colors are
  mapped between white and one random color by linearly interpolating to the
  mininum and maxinum.
- If the values are from the matrix (let's denote it as $M$) which corresponds
  to the heatmap body:
    * If the fraction of positive values in $M$ is between 25% and 75%, colors
      are mapped to blue, white and red by linearly interpolating to $-q$, 0
      and $q$, where $q$ is the maximum of $|M|$ if the number of unique
      values is less than 100, or $q$ is the 99th percentile of $|M|$. This
      color mapping is centric to zero.
    * Or else the colors are mapped to blue, white and red by linearly
      interpolating to $q_1$, $(q_1 + q_2)/2$ and $q_2$, where $q_1$ and $q_2$
      are mininum and maxinum if the number of unique values is $M$ is less
      than 100, or $q_1$ is the 1st percentile and $q_2$ is the 99th
      percentile in $M$.

`rect_gp` allows a non-standard parameter `type`. If it is set to `"none"`,
the clustering is still applied but nothing in drawn on the heatmap body. The
customized graphics on heatmap body can be added via a self-defined `cell_fun`
or `layer_fun` (see Section \@ref(customize-the-heatmap-body)).

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", rect_gp = gpar(type = "none"),
    column_title = "nothing is drawn in the heatmap body")
```

## Titles {#heatmap-titles}

The title of the heatmap basically tells what the plot is about. In
**ComplexHeatmap** package, you can set heatmap title either by the row or/and
by the column. Note at a same time you can only put e.g. column title either
on the top or at the bottom of the heatmap.

The graphic parameters can be set by `row_title_gp` and `column_title_gp`
respectively. Please remember you should use `gpar()` to specify graphic
parameters.

```{r row_column_title, fig.width = 5.2}
Heatmap(mat, name = "mat", column_title = "I am a column title", 
    row_title = "I am a row title")
Heatmap(mat, name = "mat", column_title = "I am a column title at the bottom", 
    column_title_side = "bottom")
Heatmap(mat, name = "mat", column_title = "I am a big column title", 
    column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```

Rotations for titles can be set by `row_title_rot` and `column_title_rot`, but
only horizontal and vertical rotations are allowed.

```{r title_rotation, fig.width = 5.2}
Heatmap(mat, name = "mat", row_title = "row title", row_title_rot = 0)
```

Row or column title supports as a template which is used when rows or columns
are split in the heatmap (because there will be multiple row/column titles).
This functionality is introduced in Section \@ref(heatmap-split). A quick
example would be:

```{r, eval = FALSE}
# code only for demonstration
# row title would be cluster_1 and cluster_2
Heatmap(mat, name = "mat", row_km = 2, row_title = "cluster_%s")
```

You can set `fill` parameter in `row_title_gp` and `column_title_gp` to set
the background color of titles. Since `col` in e.g. `row_title_gp` controls the 
color of text, `border` is used to control the color of the background border.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", column_title = "I am a column title", 
    column_title_gp = gpar(fill = "red", col = "white", border = "blue"))
```

You might have found there is no space on the top of the column title when the background
is colored. This can be solved by setting the global parameter `ht_opt$TITLE_PADDING`.
The use for `ht_opt()` will be introduced in Section \@ref(change-parameters-globally).

```{r, fig.width = 5.2}
ht_opt$TITLE_PADDING = unit(c(8.5, 8.5), "points")
Heatmap(mat, name = "mat", column_title = "I am a column title", 
    column_title_gp = gpar(fill = "red", col = "white", border = "blue"))
ht_opt(RESET = TRUE)
```

If the graphic elements are texts, they can be set as mathematical formulas.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", 
    column_title = expression(hat(beta) == (X^t * X)^{-1} * X^t * y)) 
```

More complicated text can be drawn as the heatmap title by using the
**gridtext** package. See Section \@ref(gridtext-title) for more examples.


## Clustering {#clustering}

Clustering might be the key component of heatmap visualization. In
**ComplexHeatmap** package, hierarchical clustering is supported with great
flexibility. You can specify the clustering either by:

- a pre-defined distance method (e.g. `"euclidean"` or `"pearson"`),
- a distance function,
- a object that already contains clustering (a `hclust` or `dendrogram` object
  or object that can be coerced to `dendrogram` class),
- a clustering function.

It is also possible to render the dendrograms with different colors and styles
for different nodes and branches for better revealing structures of the dendrogram (e.g.
by `dendextend::color_branches()`).

First, there are general settings for the clustering, e.g. whether to apply
clustering or show dendrograms, the side of the dendrograms and heights of the
dendrograms.

```{r cluster_basic, fig.width = 5.2}
Heatmap(mat, name = "mat", cluster_rows = FALSE) # turn off row clustering
Heatmap(mat, name = "mat", show_column_dend = FALSE) # hide column dendrogram
Heatmap(mat, name = "mat", row_dend_side = "right", column_dend_side = "bottom")
Heatmap(mat, name = "mat", column_dend_height = unit(4, "cm"), 
    row_dend_width = unit(4, "cm"))
```

### Distance methods {#distance-methods}

Hierarchical clustering is performed in two steps: calculate the distance
matrix and apply clustering. There are three ways to specify distance metric
for clustering:

- specify distance as a pre-defined option. The valid values are the supported
  methods in `dist()` function and in `"pearson"`, `"spearman"` and
  `"kendall"`. The correlation distance is defined as `1 - cor(x, y, method)`.
  All these built-in distance methods allow `NA` values.
- a self-defined function which calculates distance from a matrix. The
  function should only contain one argument. Please note for clustering on
  columns, the matrix will be transposed automatically.
- a self-defined function which calculates distance from two vectors. The
  function should only contain two arguments. Note this might be slow because
  it is implemented by two nested `for` loop.

```{r cluster_distance, fig.width = 5.2}
Heatmap(mat, name = "mat", clustering_distance_rows = "pearson",
    column_title = "pre-defined distance method (1 - pearson)")
Heatmap(mat, name = "mat", clustering_distance_rows = function(m) dist(m),
    column_title = "a function that calculates distance matrix")
Heatmap(mat, name = "mat", clustering_distance_rows = function(x, y) 1 - cor(x, y),
    column_title = "a function that calculates pairwise distance")
```

Based on these features, we can apply clustering which is robust to outliers
based on the pairwise distance. Note here we set the color mapping function
because we don't want outliers affect the colors.

```{r cluster_distance_advanced, eval = 1:10}
mat_with_outliers = mat
for(i in  1:10) mat_with_outliers[i, i] = 1000
robust_dist = function(x, y) {
    qx = quantile(x, c(0.1, 0.9))
    qy = quantile(y, c(0.1, 0.9))
    l = x > qx[1] & x < qx[2] & y > qy[1] & y < qy[2]
    x = x[l]
    y = y[l]
    sqrt(sum((x - y)^2))
}
```

We can compare the two heatmaps with or without the robust distance method:

```{r, eval = FALSE}
Heatmap(mat_with_outliers, name = "mat", 
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    column_title = "dist")
Heatmap(mat_with_outliers, name = "mat", 
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    clustering_distance_rows = robust_dist,
    clustering_distance_columns = robust_dist,
    column_title = "robust_dist")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
ht1 = Heatmap(mat_with_outliers, name = "mat", 
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    column_title = "dist")
draw(ht1, newpage = FALSE)
popViewport()
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
ht2 = Heatmap(mat_with_outliers, name = "mat", 
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    clustering_distance_rows = robust_dist,
    clustering_distance_columns = robust_dist,
    column_title = "robust_dist")
draw(ht2, newpage = FALSE)
popViewport()
```

If there are proper distance methods (like methods in [**stringdist**
package](https://cran.r-project.org/web/packages/stringdist/)), you can also
cluster a character matrix. `cell_fun` argument will be introduced in Section
\@ref(customize-the-heatmap-body).

```{r cluster_character_matrix, fig.width = 5.2}
mat_letters = matrix(sample(letters[1:4], 100, replace = TRUE), 10)
# distance in the ASCII table
dist_letters = function(x, y) {
    x = strtoi(charToRaw(paste(x, collapse = "")), base = 16)
    y = strtoi(charToRaw(paste(y, collapse = "")), base = 16)
    sqrt(sum((x - y)^2))
}
Heatmap(mat_letters, name = "letters", col = structure(2:5, names = letters[1:4]),
    clustering_distance_rows = dist_letters, clustering_distance_columns = dist_letters,
    cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
        grid.text(mat_letters[i, j], x, y)
    })
```

### Clustering methods {#clustering-methods}

Method to perform hierarchical clustering can be specified by
`clustering_method_rows` and `clustering_method_columns`. Possible methods are
those supported in `hclust()` function.

```{r cluster_method, fig.width = 5.2}
Heatmap(mat, name = "mat", clustering_method_rows = "single")
```

If you already have a clustering object or a function which directly returns a
clustering object, you can ignore the distance settings and set `cluster_rows`
or `cluster_columns` to the clustering objects or clustering functions. If it
is a clustering function, the only argument should be the matrix and it should
return a `hclust` or `dendrogram` object or a object that can be converted to
the `dendrogram` class.

In following example, we perform clustering with methods from **cluster**
package either by a pre-calculated clustering object or a clustering function:

```{r cluster_object, fig.width = 5.2}
library(cluster)
Heatmap(mat, name = "mat", cluster_rows = diana(mat),
   cluster_columns = agnes(t(mat)), column_title = "clustering objects")

# if cluster_columns is set as a function, you don't need to transpose the matrix
Heatmap(mat, name = "mat", cluster_rows = diana,
   cluster_columns = agnes, column_title = "clustering functions")
```

The last command is as same as :

```{r, eval = FALSE}
# code only for demonstration
Heatmap(mat, name = "mat", cluster_rows = function(m) as.dendrogram(diana(m)),
    cluster_columns = function(m) as.dendrogram(agnes(m)), 
    column_title = "clutering functions")
```

Please note, when `cluster_rows` is set as a function, the argument `m` is the
input `mat` itself, while for `cluster_columns`, `m` is the transpose of
`mat`.

`fastcluster::hclust` implements a faster version of `hclust()`. You can set
it to `cluster_rows` and `cluster_columns` to use the faster version of
`hclust()`.

```{r, eval = FALSE}
# code only for demonstration
fh = function(x) fastcluster::hclust(dist(x))
Heatmap(mat, name = "mat", cluster_rows = fh, cluster_columns = fh)
```

To make it more convinient to use the faster version of `hclust()` (assuming
you have many heatmaps to construct), it can be set as a global option. The
usage of `ht_opt` is introduced in Section
\@ref(change-parameters-globally).

```{r, eval = FALSE}
# code only for demonstration
ht_opt$fast_hclust = TRUE
# now fastcluster::hclust is used in all heatmaps
```

This is one special scenario where you might already have a subgroup
classification for the matrix rows or columns, and you only want to perform
clustering for the features in the same subgroup. You
can split the heatmap by the subgroup variable (see Section
\@ref(heatmap-split)), or you can use `cluster_within_group()` clustering
function to generate a special dendrogram.

```{r, fig.width = 5.2}
group = kmeans(t(mat), centers = 3)$cluster
Heatmap(mat, name = "mat", cluster_columns = cluster_within_group(mat, group))
```

In above example, columns in a same group are still clustered, but the
dendrogram is degenerated as a flat line. The dendrogram on columns shows the
hierarchy of the groups.

### Render dendrograms {#render-dendrograms}

If you want to render the dendrogram, normally you need to generate a
`dendrogram` object and render it through `nodePar` and `edgePar` parameter in
advance, then send it to the `cluster_rows` or `cluster_columns` argument.

You can render your `dendrogram` object by the **dendextend** package to make
a more customized visualization of the dendrogram.

```{r cluster_dendextend, fig.width = 5.2}
library(dendextend)
row_dend = as.dendrogram(hclust(dist(mat)))
row_dend = color_branches(row_dend, k = 2) # `color_branches()` returns a dendrogram object
Heatmap(mat, name = "mat", cluster_rows = row_dend)
```

`row_dend_gp` and `column_dend_gp` control the global graphic setting for
dendrograms. Note e.g. graphic settings in `row_dend` will be overwritten by
`row_dend_gp`.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", cluster_rows = row_dend, row_dend_gp = gpar(col = "red"))
```

From version 2.5.6, you can also add graphics on the nodes of the dendrogram by setting
a proper `nodePar`.

```{r, fig.width = 5.2}
row_dend = dendrapply(row_dend, function(d) {
    attr(d, "nodePar") = list(cex = 0.8, pch = sample(20, 1), col = rand_color(1))
    return(d)
})
Heatmap(mat, name = "mat", cluster_rows = row_dend, row_dend_width = unit(2, "cm"))
```

Currently, only points can be added on the nodes, however, if you have the need to add
more types of graphics, especially the self-defined graphics, just send me a message and I
will think about it.

### Reorder dendrograms {#reorder-dendrograms}

In the `Heatmap()` function, dendrograms are reordered to make features with
larger difference more separated from each others (please refer to the
documentation of `reorder.dendrogram()`). Here the difference (or it is called
the weight) is measured by the row means if it is a row dendrogram or by the
column means if it is a column dendrogram. `row_dend_reorder` and
`column_dend_reorder` control whether to apply dendrogram reordering if the
value is set as logical. The two arguments also control the weight for the
reordering if they are set to numeric vectors (it will be sent to the `wts`
argument of `reorder.dendrogram()`). The reordering can be turned off by
setting e.g. `row_dend_reorder = FALSE`.

By default, dendrogram reordering is turned on if
`cluster_rows`/`cluster_columns` is set as logical value or a clustering
function. It is turned off if `cluster_rows`/`cluster_columns` is set as
clustering object.

Compare following two heatmaps:

```{r, eval = FALSE}
m2 = matrix(1:100, nr = 10, byrow = TRUE)
Heatmap(m2, name = "mat", row_dend_reorder = FALSE, column_title = "no reordering")
Heatmap(m2, name = "mat", row_dend_reorder = TRUE, column_title = "apply reordering")
```

```{r cluster_dendsort, fig.width = 10, fig.height = 5, echo = FALSE}
m2 = matrix(1:100, nr = 10, byrow = TRUE)
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(Heatmap(m2, name = "mat", row_dend_reorder = FALSE, column_title = "no reordering"), 
    newpage = FALSE)
upViewport()

pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(Heatmap(m2, name = "mat", row_dend_reorder = TRUE, column_title = "applied reordering"), 
    newpage = FALSE)
upViewport()
```

There are many other methods for reordering dendrograms, e.g. the **dendsort**
package. Basically, all these methods still return a dendrogram that has been
reordered, thus, we can firstly generate the row or column dendrogram based on
the data matrix, reorder it by some method, and assign it back to
`cluster_rows` or `cluster_columns`.

Compare following two reorderings on both dimensions. Can you tell which is better?

```{r, eval = FALSE}
Heatmap(mat, name = "mat", column_title = "default reordering")

library(dendsort)
row_dend = dendsort(hclust(dist(mat)))
col_dend = dendsort(hclust(dist(t(mat))))
Heatmap(mat, name = "mat", cluster_rows = row_dend, cluster_columns = col_dend,
    column_title = "reorder by dendsort")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(Heatmap(mat, name = "mat", column_title = "default reordering"), newpage = FALSE)
popViewport()

library(dendsort)
dend = dendsort(hclust(dist(mat)))
col_dend = dendsort(hclust(dist(t(mat))))
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(Heatmap(mat, name = "mat", cluster_rows = row_dend, cluster_columns = col_dend,
    column_title = "reordering by dendsort"), newpage = FALSE)
popViewport()
```

## Set row and column orders {#row-and_column_orders}

Clustering is used to adjust row orders and column orders of the heatmap, but
you can still set the order manually by `row_order` and `column_order`. If
e.g. `row_order` is set, row clustering is turned off by default.

```{r manual_order, fig.width = 5.2}
Heatmap(mat, name = "mat", row_order = order(as.numeric(gsub("row", "", rownames(mat)))), 
    column_order = order(as.numeric(gsub("column", "", colnames(mat)))),
    column_title = "reorder matrix")
```

The orders can be character vectors if they are just shuffles of the matrix row names or column names.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_order = sort(rownames(mat)), 
    column_order = sort(colnames(mat)),
    column_title = "reorder matrix by row/column names")
```

Note `row_dend_reorder` and `row_order` are two different things.
`row_dend_reorder` is applied on the dendrogram. For any node in the
dendrogram, rotating its two branches actually gives an identical dendrogram,
thus, reordering the dendrogram by automatically rotating sub-dendrogram at
every node can help to separate features further from each other which show
more difference. As a comparison, `row_order` is simply applied on the matrix
and normally dendrograms should be turned off.

## Seriation {#heatmap-seriation}

Seriation is an interesting technique for ordering the matrix (see this
interesting post: http://nicolas.kruchten.com/content/2018/02/seriation/). The
powerful [**seriation**
package](https://cran.r-project.org/web/packages/seriation/index.html)
implements quite a lot of methods for seriation. Since it is easy to extract
row orders and column orders from the object returned by the core function
`seriate()` from **seriation** package, they can be directly assigned to
`row_order` and `column_order` to make the heatmap.

The first example demonstrates to directly apply `seriate()` on the matrix.
Since the `"BEA_TSP"` method only allows a non-negative matrix, we modify the
matrix to `max(mat) - mat`.

```{r, fig.width = 5.2}
library(seriation)
o = seriate(max(mat) - mat, method = "BEA_TSP")
Heatmap(max(mat) - mat, name = "mat", 
    row_order = get_order(o, 1), column_order = get_order(o, 2),
    column_title = "seriation by BEA_TSP method")
```

Or you can apply `seriate()` to the distance matrix. Now the order for rows
and columns needs to be calcualted separatedly because the distance matrix
needs to be calculated separatedly for columns and rows.

```{r, fig.width = 5.2}
o1 = seriate(dist(mat), method = "TSP")
o2 = seriate(dist(t(mat)), method = "TSP")
Heatmap(mat, name = "mat", row_order = get_order(o1), column_order = get_order(o2),
    column_title = "seriation from the distance matrix")
```

Some seriation methods also contain the hierarchical clustering information.
Let's try:

```{r}
o1 = seriate(dist(mat), method = "GW")
o2 = seriate(dist(t(mat)), method = "GW")
```

`o1` and `o2` are actually mainly composed of `hclust` objects:

```{r}
class(o1[[1]])
```

And the orders are the same by using `hclust$order` or `get_order()`.

```{r}
o1[[1]]$order
# should be the same as the previous one
get_order(o1)
```

And we can add the dendrograms to the heatmap.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", cluster_rows = as.dendrogram(o1[[1]]), 
    cluster_columns = as.dendrogram(o2[[1]]))
```

For more use of the `seriate()` function, please refer to the [**seriation**
package](https://cran.r-project.org/web/packages/seriation/index.html).

## Dimension names {#dimension-names}

The row names and column names are drawn on the right and bottom sides of the
heatmap by default. Side, visibility and graphic parameters for dimension
names can be set as follows:

```{r dimension_name, fig.width = 5.2}
Heatmap(mat, name = "mat", row_names_side = "left", row_dend_side = "right", 
    column_names_side = "top", column_dend_side = "bottom")
Heatmap(mat, name = "mat", show_row_names = FALSE)
Heatmap(mat, name = "mat", row_names_gp = gpar(fontsize = 20))
Heatmap(mat, name = "mat", row_names_gp = gpar(col = c(rep("red", 10), rep("blue", 8))))
Heatmap(mat, name = "mat", row_names_centered = TRUE, column_names_centered = TRUE)
```

The rotation of column names can be set by `column_names_rot`:

```{r, eval = FALSE}
Heatmap(mat, name = "mat", column_names_rot = 45)
Heatmap(mat, name = "mat", column_names_rot = 45, column_names_side = "top",
    column_dend_side = "bottom")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(Heatmap(mat, name = "mat", column_names_rot = 45), newpage = FALSE)
upViewport()
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(Heatmap(mat, name = "mat", column_names_rot = 45, column_names_side = "top",
    column_dend_side = "bottom"), newpage = FALSE)
upViewport()
```

If you have row names or column names which are too long,
`row_names_max_width` or `column_names_max_height` can be used to set the
maximal space for them. The default maximal space for row names and column
names are all 6 cm. In following code, `max_text_width()` is a helper function
to quick calculate maximal width from a vector of text.

```{r, eval = FALSE}
mat2 = mat
rownames(mat2)[1] = paste(c(letters, LETTERS), collapse = "")
Heatmap(mat2, name = "mat", row_title = "default row_names_max_width")
Heatmap(mat2, name = "mat", row_title = "row_names_max_width as length of a*",
    row_names_max_width = max_text_width(
        rownames(mat2), 
        gp = gpar(fontsize = 12)
    ))
```

```{r, echo = FALSE, fig.width = 7, fig.height = 8}
mat2 = mat
rownames(mat2)[1] = paste(c(letters, LETTERS), collapse = "")
pushViewport(viewport(y = 1, height = 0.5, just = "top"))
draw(Heatmap(mat2, name = "mat", row_title = "default row_names_max_width"), newpage = FALSE)
upViewport()
pushViewport(viewport(y = 0.5, height = 0.5, just = "top"))
draw(Heatmap(mat2, name = "mat", row_title = "row_names_max_width as length of a*",
    row_names_max_width = max_text_width(rownames(mat2), gp = gpar(fontsize = 12))), newpage = FALSE)
upViewport()
```

Instead of directly using the row/column names from the matrix, you can also
provide another character vector which corresponds to the rows or columns and
set it by `row_labels` or `column_labels`. This is useful because you don't
need to change the dimension names of the matrix to change the labels on the
heatmap while you can directly provide the new labels.

There is one typical scenario that `row_labels` and `column_labels` are
useful. For the gene expression analysis, we might use Ensembl ID as the gene
ID which is used as row names of the gene expression matrix. However, the
Ensembl ID is for the indexing of the Ensembl database but not for the human
reading. Instead, we would prefer to put gene symbols on the heatmap as the
row names which is easier to read. To do this, we only need to assign the
corresponding gene symbols to `row_labels` without modifying the original
matrix.

The second advantage is `row_labels` or `column_labels` allows duplicated
labels, while duplicated row names or column names are not allowed in the
matrix.

Following gives a simple example that we put letters as row labels and column
labels:

```{r, fig.width = 5.2}
# use a named vector to make sure the correspondance between 
# row names and row labels is correct
row_labels = structure(paste0(letters[1:24], 1:24), names = paste0("row", 1:24))
column_labels = structure(paste0(LETTERS[1:24], 1:24), names = paste0("column", 1:24))
row_labels
Heatmap(mat, name = "mat", row_labels = row_labels[rownames(mat)], 
    column_labels = column_labels[colnames(mat)])
```

The third advantage is mathematical expression can be used as row names in the
heatmap.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_labels = expression(alpha, beta, gamma, delta, epsilon, 
    zeta, eta, theta, iota, kappa, lambda, mu, nu, xi, omicron, pi, rho, sigma))
```

Internally, row names and columns names are actually implemented by the
`anno_text()` function (Section \@ref(text-annotation)), in other words, they
are treated as special cases for text annotations.

Complicated styles of row/column names can be implemented by integrating the
**gridtext** package. Check Section \@ref(gridtext-dimnames) for more examples.

## Heatmap split {#heatmap-split}

One major advantage of **ComplexHeatmap** package is that it supports splitting the
heatmap by rows and columns to better group the features and additionally
highlight the patterns.

Following arguments control the splitting: `row_km`, `row_split`, `column_km`,
`column_split`. In following, we call the sub-clusters generated by splitting as
"_slices_".

### Split by k-means clustering {#split-by-kmeans-clustering}

`row_km` and `column_km` apply k-means partitioning.

```{r k_means, fig.width = 5.2}
Heatmap(mat, name = "mat", row_km = 2)
Heatmap(mat, name = "mat", column_km = 3)
```

Row splitting and column splitting can be performed simultaneously.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_km = 2, column_km = 3)
```

You might notice there are dashed lines in the row and column dendrograms,
it will be explained in Section \@ref(split-by-categorical-variables) (the last paragraph).

`Heatmap()` internally calls `kmeans()` with random start points, which
results in, for some cases, generating different clusters from repeated runs.
To get rid of this problem, `row_km_repeats` and `column_km_repeats` can be
set to a number larger than 1 to run `kmeans()` multiple times and a final
consensus k-means clustering is used. Please note the final number of clusters
form consensus k-means might be smaller than the number set in `row_km` and
`column_km`.

```{r, eval = FALSE}
# of course, it will be a little bit slower
Heatmap(mat, name = "mat", 
    row_km = 2, row_km_repeats = 100,
    column_km = 3, column_km_repeats = 100)
```

### Split by categorical variables {#split-by-categorical-variables}

More generally, `row_split` or `column_split` can be set to a categorical
vector or a data frame where different combinations of levels split the
rows/columns in the heatmap. How to control the order of the slices is
introduced in Section \@ref(order-of-slices).

```{r split, fig.width = 5.2}
# split by a vector
Heatmap(mat, name = "mat", 
    row_split = rep(c("A", "B"), 9), column_split = rep(c("C", "D"), 12))
# split by a data frame
Heatmap(mat, name = "mat", 
    row_split = data.frame(rep(c("A", "B"), 9), rep(c("C", "D"), each = 9)))
# split on both dimensions
Heatmap(mat, name = "mat", row_split = factor(rep(c("A", "B"), 9)),
    column_split = factor(rep(c("C", "D"), 12)))
```

Actually, k-means clustering just generates a vector of cluster classes and
appends to `row_split` or `column_split`. `row_km`/`column_km` can be used
mixed with `row_split` and `column_split`.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_split = rep(c("A", "B"), 9), row_km = 2)
```

which is the same as:

```{r, eval = FALSE}
# code only for demonstration
cl = kmeans(mat, centers = 2)$cluster
# classes from k-means are always put as the first column in `row_split`
Heatmap(mat, name = "mat", row_split = cbind(cl, rep(c("A", "B"), 9)))
```

If you are not happy with the default k-means partition, it is easy to use
other partition methods by just assigning the partition vector to
`row_split`/`column_split`.

```{r pam, fig.width = 5.2}
pa = cluster::pam(mat, k = 3)
Heatmap(mat, name = "mat", row_split = paste0("pam", pa$clustering))
```

If `row_order` or `column_order` is set, in each row/column slice, it is still
ordered.

```{r split_row_order, fig.width = 5.2}
# remember when `row_order` is set, row clustering is turned off
Heatmap(mat, name = "mat", row_order = 18:1, row_km = 2)
```

Character matrix can only be split by `row_split`/`column_split` argument.

```{r split_discrete_matrix, fig.width = 5.2}
# split by the first column in `discrete_mat`
Heatmap(discrete_mat, name = "mat", col = 1:4, row_split = discrete_mat[, 1])
```

If `row_km`/`column_km` is set or `row_split`/`column_split` is set as a
vector or a data frame, hierarchical clustering is first applied to each slice
which generates `k` dendrograms, then a parent dendrogram is generated based
on the mean values of each slice. **The height of the parent dendrogram is
adjusted by adding the maximal height of the dendrograms in all children
slices and the parent dendrogram is added on top of the children dendrograms
to form a single global dendrogram.** This is why you see dashed lines in the
dendrograms in previous heatmaps. They are used to discriminate the parent dendrogram
and the children dendrograms, and alert users they are calculated in different
ways. These dashed lines can be removed by setting `show_parent_dend_line =
FALSE` in `Heatmap()`, or set it as a global option:
`ht_opt$show_parent_dend_line = FALSE`.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_km = 2, column_km = 3, show_parent_dend_line = FALSE)
```

### Split by dendrogram {#spilt-by-dendrogram}

A second scenario for splitting is that users may still want to keep the
global dendrogram **which is generated from the complete matrix** while not
split it in the first place. In this case, `row_split`/`column_split` can be
set to a single number which will apply `cutree()` on the row/column
dendrogram. This works when `cluster_rows`/`cluster_columns` is set to `TRUE`
or is assigned with a `hclust`/`dendrogram` object.

For this case, the dendrogram is still as same as the original one, except that the
positions of dendrogram leaves are slightly adjusted by the gaps between
slices. (There is no dashed lines, because here the dendrogram is calcualted
as a complete one and there is no parent dendrogram or children dendrograms.)

```{r split_dendrogram, fig.width = 5.2}
Heatmap(mat, name = "mat", row_split = 2, column_split = 3)
dend = hclust(dist(mat))
dend = color_branches(dend, k = 2)
Heatmap(mat, name = "mat", cluster_rows = dend, row_split = 2)
```

If you want to combine splitting from `cutree()` and other categorical
variables, you need to generate the classes from `cutree()` in the first
place, append to e.g. `row_split` as a data frame and then send it to
`row_split` argument.

```{r, eval = FALSE}
# code only for demonstration
split = data.frame(cutree(hclust(dist(mat)), k = 2), rep(c("A", "B"), 9))
Heatmap(mat, name = "mat", row_split = split)
```

### Order of slices {#order-of-slices}

When `row_split`/`column_split` is set as a categorical variable (a vector or a
data frame) or `row_km`/`column_km` is set, by default, there is an additional
clustering applied to the mean of slices to show the hierarchy in the slice
level. Under this scenario, you cannot precisely control the order of slices
because it is controlled by the clustering of slices.

Nevertheless, you can set `cluster_row_slices` or `cluster_column_slices` to
`FALSE` to turn off the clustering on slices, and now you can precisely
control the order of slices.

When there is no slice clustering, the order of each slice can be controlled
by `levels` of each variable in `row_split`/`column_split` (in this case, each
variable should be a factor). If all variables are characters, the default
order is `unique(row_split)` or `unique(column_split)`. Compare following
heatmaps:

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", 
  row_split = rep(LETTERS[1:3], 6),
    column_split = rep(letters[1:6], 4))

# clustering is similar as previous heatmap with branches in some nodes in the dendrogram flipped
Heatmap(mat, name = "mat", 
  row_split = factor(rep(LETTERS[1:3], 6), levels = LETTERS[3:1]),
    column_split = factor(rep(letters[1:6], 4), levels = letters[6:1]))

# now the order is exactly what we set
Heatmap(mat, name = "mat", 
  row_split = factor(rep(LETTERS[1:3], 6), levels = LETTERS[3:1]),
    column_split = factor(rep(letters[1:6], 4), levels = letters[6:1]),
    cluster_row_slices = FALSE, 
    cluster_column_slices = FALSE)
```


### Titles for splitting {#titles-for-splitting} 

When `row_split`/`column_split` is set as a single number, there is only one
categorical variable, while when `row_km`/`column_km` is set and/or
`row_split`/`column_split` is set as categorical variables, there will be
multiple categorical variables. By default, the titles are in a form of
`"level1,level2,..."` which corresponds to every combination of levels in all
categorical variables. The titles for splitting can be controlled by "a
template".

**ComplexHeatmap** supports three types of templates. The first one is by
`sprintf()` where the `%s` is replaced by the corresponding level. In
following example, since all combinations of `split` are `A,C`, `A,D`, `B,C`
and `B,D`, if `row_title` is set to `%s|%s`, the four row titles will be
`A|C`, `A|D`, `B|C`, `B|D`.

```{r, fig.width = 5.2}
split = data.frame(rep(c("A", "B"), 9), rep(c("C", "D"), each = 9))
Heatmap(mat, name = "mat", row_split = split, row_title = "%s|%s")
```

For the `sprintf()` template, you can only put the levels which are `A,B,C,D`
in the title, and `C,D` is always after `A,B` (i.e. it always `A,C` and
`A,D`). However, when making the heatmap, you might want to put more
meaningful text instead of the internal levels. Once you know how to
correspond the text to the level, you can add it by following two template
methods.

In the following two template methods, special marks are used to mark the R
code which is executable (it is called variable interpolation where the code
is extracted and executed and the returned value in put back to the string).
There are two types of template marks `@{}` and `{}`. The first one is from
**GetoptLong** package which should already be installed when you install the
**ComplexHeatmap** package and the second one is from **glue** package which
you need to install first.

There is an internal variable `x` you should use when you use the latter two
templates. `x` is just a simple vector which contains current category levels
(e.g. `c("A", "C")`).

```{r, eval = FALSE}
# We only run the code for the first heatmap
map = c("A" = "aaa", "B" = "bbb", "C" = "333", "D" = "444")
Heatmap(mat, name = "mat", row_split = split, row_title = "@{map[ x[1] ]}|@{map[ x[2] ]}")
Heatmap(mat, name = "mat", row_split = split, row_title = "{map[ x[1] ]}|{map[ x[2] ]}")
```

```{r, echo = FALSE, fig.width = 5.2}
map = c("A" = "aaa", "B" = "bbb", "C" = "333", "D" = "444")
Heatmap(mat, name = "mat", row_split = split, row_title = "@{map[ x[1] ]}|@{map[ x[2] ]}")
```

The row title is rotated by default, you can set `row_title_rot = 0` to make
it horizontal:

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_split = split, row_title = "%s|%s", row_title_rot = 0)
```

When `row_split`/`column_split` is set as a number, you can also use template
to adjust the titles for slices.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_split = 2, row_title = "cluster_%s")
```

If you know the final number of row slices, you can directly set a vector of
titles to `row_title`. Be careful the number of row slices is not always
identical to `nlevel_1*nlevel_2*...`.

```{r}
# we know there are four slices
Heatmap(mat, name = "mat", row_split = split, 
    row_title = c("top_slice", "middle_top_slice", "middle_bottom_slice", "bottom_slice"),
    row_title_rot = 0)
```

If the length of `row_title` is specified as a single string, it will be like
a single title for all slices.


```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_split = split, row_title = "there are four slices")
```

If you still want titles for each slice, but also a global title, you can do
as follows.

```{r, fig.width = 5.2}
ht = Heatmap(mat, name = "mat", row_split = split, row_title = "%s|%s")
# This row_title is actually a heatmap-list-level row title
draw(ht, row_title = "I am a row title")
```

Actually the `row_title` used in `draw()` function is the row title of the
heatmap list, although in the example there is only one heatmap. The `draw()`
function and the heatmap list will be introduced in Chapter
\@ref(a-list-of-heatmaps).

If `row_title` is set to `NULL`, no row title is drawn.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_split = split, row_title = NULL)
```

All these rules also work for column titles for slices.

### Graphic parameters for splitting {#graphic-parameters-for-splitting}

When splitting is applied on rows/columns, graphic parameters for row/column
title and row/column names can be specified as same length as number of
slices.

```{r split_graphical_parameter, fig.width = 5.2}
# by defalt, there no space on the top of the title, here we add 4pt to the top.
# it can be reset by `ht_opt(RESET = TRUE)`
ht_opt$TITLE_PADDING = unit(c(4, 4), "points")
Heatmap(mat, name = "mat", 
    row_km = 2, row_title_gp = gpar(col = c("red", "blue"), font = 1:2),
    row_names_gp = gpar(col = c("green", "orange"), fontsize = c(10, 14)),
    column_km = 3, column_title_gp = gpar(fill = c("red", "blue", "green"), font = 1:3),
    column_names_gp = gpar(col = c("green", "orange", "purple"), fontsize = c(10, 14, 8)))
```

```{r, echo = FALSE}
ht_opt(RESET = TRUE)
```

### Gaps between slices {#gaps-between-slices}

The space of gaps between row/column slices can be controlled by
`row_gap`/`column_gap`. The value can be a single unit or a vector of units.

```{r split_gap, fig.width = 5.2}
Heatmap(mat, name = "mat", row_km = 3, row_gap = unit(5, "mm"))
Heatmap(mat, name = "mat", row_km = 3, row_gap = unit(c(2, 4), "mm"))
Heatmap(mat, name = "mat", row_km = 3, row_gap = unit(c(2, 4), "mm"),
    column_km = 3, column_gap = unit(c(2, 4), "mm"))
```

When heatmap border is added by setting `border = TRUE`, the border of every
slice is added.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_km = 2, column_km = 3, border = TRUE)
```

If you set gap size to zero, the heatmap will look like it is partitioned by
vertical and horizontal lines.

```{r, fig.width = 5.2}
Heatmap(mat, name = "mat", row_km = 2, column_km = 3, 
    row_gap = unit(0, "mm"), column_gap = unit(0, "mm"), border = TRUE)
```

### Split heatmap annotations {#split-heatmap-annotations}

When the heatmap is split, all the heatmap components are split accordingly.
Following gives you a simple example and the heatmap annotation will be
introduced in Chapter \@ref(heatmap-annotations).

```{r, fig.width = 6}
Heatmap(mat, name = "mat", row_km = 2, column_km = 3,
    top_annotation = HeatmapAnnotation(foo1 = 1:24, bar1 = anno_points(runif(24))),
    right_annotation = rowAnnotation(foo2 = 18:1, bar2 = anno_barplot(runif(18)))
)
```

## Heatmap as raster image {#heatmap-as-raster-image}

When we produce so-called "high quality figures", normally we save the figures
as [vector graphics](https://en.wikipedia.org/wiki/Vector_graphics) in the
format of _e.g._ pdf or svg. The vector graphics basically store details of
every single graphic elements, thus, if a heatmap made from a very huge matrix is
saved as vector graphics, the final file size would be very big. On the other
hand, when visualizing _e.g._ the pdf file on the screen, multiple grids from
the heatmap actually only map to single pixels, due to the limited size of the
screen. Thus, there need some ways to effectively reduce the original image and
it is not necessary to store the complete matrix for the heatmap.

Rasterization is a way to covnert the vector graphics into a matrix of colors.
In this case, an image is represented as a matrix of RGB values, which is
called a [raster image](https://en.wikipedia.org/wiki/Raster_graphics). If the
heatmap is larger than the size of the screen or the pixels that current
graphics devices can support, we can convert the heatmap and reduce it, by
saving it in a form of a color matrix with the same dimension as the device.

Let's assume a matrix has $n_r$ rows and $n_c$ columns. When it is drawn on a
certain graphics device, _e.g._ an on-screen device, the corresponding heatmap
body has $p_r$ and $p_c$ pixels (or points) for the rows and columns,
respectively. When $n_r > p_r$ and/or $n_c > p_c$, multiple values in the
matrix are mapped to single pixels. Here we need to reduce $n_r$ and/or $n_c$
if they are larger than $p_r$ and/or $p_c$.

To make it simple, I assume both $n_r > p_r$ and $n_c > p_c$. The
principle is basically the same for the scenarios where only one dimension of
the matrix is larger than the device.

In **ComplexHeatmap** version 2.5.4, there are following implementations for
image rasterization. Note the implementation is a little bit different from
the earlier versions (of course, better than the earlier versions).

1. First an image
   (in a specific format, _e.g._ png or jpeg) with $(p_r \cdot a) \times (p_c
   \cdot a)$ resolution is saved into a temporary file where $a$ is a zooming
   factor, next it is read back as a `raster` object by _e.g._
   `png::readPNG()` or `jpeg::readJPEG()`, and later the raster object is
   filled into the heatmap body by `grid::grid.raster()`. So we can say, the 
   rasterization is done by the raster image devices (`png()` or `jpeg()`).

   This type of rasterization is automatically turned on (**if magick package is
   not installed**) when the number of rows or columns exceeds 2000 (You will
   see a message. It won't happen silently). It can also be manually
   controlled by setting the `use_raster` argument:

```{r, eval = FALSE, echo = TRUE}
Heatmap(..., use_raster = TRUE)
```

   The zooming factor is controlled by `raster_quality` argument. A value larger
   than 1 generates files with larger size.

```{r, eval = FALSE, echo = TRUE}
Heatmap(..., use_raster = TRUE, raster_quality = 5)
```

2. Simply reduce the original matrix to $p_r \times p_c$ where now each single values can
   correspond to single pixels. In the reduction, a user-defined function is applied to summarize
   the sub-matrices.

   This can be set by `raster_resize_mat` argument:

```{r, eval = FALSE, echo = TRUE}
# the default summary function is mean()
Heatmap(..., use_raster = TRUE, raster_resize_mat = TRUE)
# use max() as the summary function
Heatmap(..., use_raster = TRUE, raster_resize_mat = max)
# randomly pick one
Heatmap(..., use_raster = TRUE, raster_resize_mat = function(x) sample(x, 1))
```

3. A temporary image with resolution $n_r \times n_c$ is first generated, here
   `magick::image_resize()` is used to reduce the image to size $p_r \times
   p_c$. Finally the reduced image is read as a `raster` object and filled into
   the heatmap body. **magick** provides a lot of methods for
   "resizing"/"scaling" the image, which is called the "filtering methods"
   under the term of **magick**. [All filtering
   methods](https://www.imagemagick.org/Magick++/Enumerations.html#FilterTypes)
   can be obtained by `magick::filter_types()`.

   This type of rasterization can be truned on by setting `raster_by_magick = TRUE` 
   and choosing a proper `raster_magick_filter`.

```{r, eval = FALSE, echo = TRUE}
Heatmap(..., use_raster = TRUE, raster_by_magick = TRUE)
Heatmap(..., use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter = ...)
```

The type of temporary image is controlled by `raster_device` argument. All supported
"raster devices" are: `png`, `CairoPNG`, `agg_png`, `jpeg`, `tiff`, `CairoJPEG` and 
`CairoTIFF`. `CairoPNG` is taken as the default raster device because the previously
default device `png` occasionally produces white vertical and horizontal lines in the
raster image.

In **ComplexHeatmap**, `use_raster` is by default turned on if the number of
rows or columns is more than 2000 in the matrix.

In the following parts of this post, I compare the visual difference
between different image rasterization methods.


The following example is from [Guillaume Devailly's simulated
data](https://gdevailly.netlify.app/post/plotting-big-matrices-in-r/) but with
small adaptation. This example shows an enrichment pattern to the top center of the plot.

```{r, echo = FALSE}
library(RColorBrewer)
mat2 = matrix(nrow = 5000, ncol = 50)
for(i in 1:5000) {
    mat2[i, ] = runif(50) + c(sort(abs(rnorm(50)))[1:25], rev(sort(abs(rnorm(50)))[1:25]))  * i/1000
}
mat2 = mat2[nrow(mat2):1, ]
col_fun2 = colorRamp2(seq(quantile(mat2, 0.01), quantile(mat2, 0.99), len = 11), rev(brewer.pal(11, "Spectral")))
```

In the folowing examples, I won't show the code for making heatmaps because there are too many heatmaps
and the specific settings is already written as the row title of each heatmap.

I set the same color mapping for all heatmaps, so that you can see how different rasterizations change
the original patterns.

For the comparison, I generated many heatmaps. They can be categoried into three groups, as corresponded to the 
three rasterization methods I mentioned previously.

- by `png()`/`CairoPNG()`/`agg_png()`/`jpeg()`/`tiff()`: Rasterization method 1. First row in the following heatmaps.
- `raster_resize_mat = *`: Rasterization method 2, with different summary methods. Second row in the following heatmaps.
- `filter = *`: Rasterization method 3, with different filterring method. The string `filter`
  should be `raster_magick_filter`. It is truncated so that the row title won't be cut by the plot regions.

```{r, fig.show = "hold", echo = FALSE, fig.width = 1.6*5, fig.height = 1.6, out.width = "100%", fig.align = "left"}
library(ComplexHeatmap)
library(circlize)
library(magick)
library(GetoptLong)
library(cowplot)
pl = list(
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_quality = 1, raster_device = "png",
        row_title = "by png()", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_quality = 1, raster_device = "CairoPNG",
        row_title = "by CairoPNG()", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_quality = 1, raster_device = "agg_png",
        row_title = "by agg_png()", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_quality = 1, raster_device = "jpeg",
        row_title = "by jpeg()", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_quality = 1, raster_device = "tiff",
        row_title = "by tiff()", row_title_gp = gpar(fontsize = 8))))
)
cowplot::plot_grid(plotlist = pl, ncol = 5)
```

```{r, fig.show = "hold", echo = FALSE, fig.width = 1.6*5, fig.height = 1.6, out.width = "100%", fig.align = "left"}
random = function(x) sample(x, 1)
first = function(x) x[1]
pl = list(
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_resize_mat = mean, 
        row_title = "raster_resize_mat = mean", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_resize_mat = max, 
        row_title = "raster_resize_mat = max", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_resize_mat = min, 
        row_title = "raster_resize_mat = min", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_resize_mat = random, 
        row_title = "raster_resize_mat = random", row_title_gp = gpar(fontsize = 8)))),
    grid.grabExpr(draw(Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = FALSE, raster_resize_mat = first, 
        row_title = "raster_resize_mat = first", row_title_gp = gpar(fontsize = 8))))
)
cowplot::plot_grid(plotlist = pl, ncol = 5)
```

```{r, fig.show = "hold", echo = FALSE, fig.width = 1.6*5, fig.height = 1.6*7, out.width = "100%", fig.align = "left"}
pl = list()
for(filter in filter_types()) {
    ht = Heatmap(mat2, col = col_fun2, cluster_rows = FALSE, cluster_columns = FALSE, show_heatmap_legend = FALSE,
        use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter = filter,
        row_title = qq("filter = '@{filter}'"), row_title_gp = gpar(fontsize = 8))
    pl[[filter]] = grid.grabExpr(draw(ht))
}
cowplot::plot_grid(plotlist = pl, ncol = 5)
```

More examples on image rasteration can be found in this blog post: [Rasterization in ComplexHeatmap
](https://jokergoo.github.io/2020/06/30/rasterization-in-complexheatmap/).

According to the examples that have been shown, I would say
rasterization by **magick** package performs better, thus, by default, in
**ComplexHeatmap**, the rasterization is done by **magick** (with `"Lanczos"`
as the default filter method) and if **magick** is not installed, it uses
`CairoPNG()` and a friendly message is printed to suggest users to install
**magick**.

Following example compares the PDF file size with raster images by different
raster devices.

```{r, results = "hide"}
library(GetoptLong)
set.seed(123)
mat2 = matrix(rnorm(10000*100), ncol = 100)
pdf(qq("heatmap_no_rasteration.pdf"), width = 8, height = 8)
ht = Heatmap(mat2, cluster_rows = FALSE, cluster_columns = FALSE, use_raster = FALSE)
draw(ht)
dev.off()

# Here I removed CairoTIFF because it is not working on my laptop
all_devices = c("png", "CairoPNG", "agg_png", "jpeg", "CairoJPEG", "tiff")
for(device in all_devices) {
    pdf(qq("heatmap_@{device}.pdf"), width = 8, height = 8)
    ht = Heatmap(mat2, cluster_rows = FALSE, cluster_columns = FALSE, use_raster = TRUE, 
        raster_device = device)
    draw(ht)
    dev.off()
}
```

```{r}
all_files = qq("heatmap_@{all_devices}.pdf", collapse = FALSE)
all_files = c("heatmap_no_rasteration.pdf", all_files)
fs = file.size(all_files)
names(fs) = all_files
sapply(fs, function(x) paste(round(x/1024), "KB"))
```

```{r, echo = FALSE, results = "hide"}
file.remove(all_files)
```

`CairoPNG` generates smaller file size, thus it is used as the default raster device.

## Customize the heatmap body {#customize-the-heatmap-body}

The heatmap body can be self-defined to add more types of graphics. By default
the heatmap body is composed by a matrix of small rectangles (it might be
called grids in other parts of this documentation, but let's call it "_cells_"
here) with different filled colors. However, it is also possible to add more
graphics or symbols as additional layers on the heatmap. There are two
arguments `cell_fun` and `layer_fun` which both should be user-defined
functions.

### cell_fun {#cell-fun}

`cell_fun` draws in each cell repeatedly, which is internally executed in two
nested `for` loops, while `layer_fun` is the vectorized version of `cell_fun`.
`cell_fun` is easier to understand but `layer_fun` is much faster to execute
and more customizable.

`cell_fun` expects a function with 7 arguments (the argument names can be
different from following, but the order must be the same), which are:

- `j`: column index in the matrix. Column index corresponds to the x-direction
  in the viewport, that's why `j` is put as the first argument.
- `i`: row index in the matrix.
- `x`: x coordinate of middle point of the cell which is measured in the
  viewport of the heatmap body.
- `y`: y coordinate of middle point of the cell which is measured in the
  viewport of the heatmap body.
- `width`: width of the cell. The value is `unit(1/ncol(sub_mat), "npc")`
  where `sub_mat` correspond to the sub-matrix by row splitting and column
  splitting.
- `height`: height of the cell. The value is `unit(1/nrow(sub_mat), "npc")`.
- `fill`: color of the cell.

The values for the seven arguments are automatically sent to the function when
executed in each cell.

The most common use is to add values in the matrix onto the heatmap:

```{r, fig.width = 5.2}
small_mat = mat[1:9, 1:9]
col_fun = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
Heatmap(small_mat, name = "mat", col = col_fun,
    cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", small_mat[i, j]), x, y, gp = gpar(fontsize = 10))
})
```

and we can also choose only to add text for the cells with positive values:

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat",  col = col_fun,
    cell_fun = function(j, i, x, y, width, height, fill) {
        if(small_mat[i, j] > 0)
            grid.text(sprintf("%.1f", small_mat[i, j]), x, y, gp = gpar(fontsize = 10))
})
```

You can split the heatmap without doing anything extra to `cell_fun`:

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat", col = col_fun,
    row_km = 2, column_km = 2,
    cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", small_mat[i, j]), x, y, gp = gpar(fontsize = 10))
})
```

In following example, we make a heatmap which shows correlation matrix similar
as the **corrplot** package:

```{r cell_fun, fig.width = 6.5, fig.height = 6}
cor_mat = cor(small_mat)
od = hclust(dist(cor_mat))$order
cor_mat = cor_mat[od, od]
nm = rownames(cor_mat)
col_fun = circlize::colorRamp2(c(-1, 0, 1), c("green", "white", "red"))
# `col = col_fun` here is used to generate the legend
Heatmap(cor_mat, name = "correlation", col = col_fun, rect_gp = gpar(type = "none"), 
    cell_fun = function(j, i, x, y, width, height, fill) {
        grid.rect(x = x, y = y, width = width, height = height, 
            gp = gpar(col = "grey", fill = NA))
        if(i == j) {
            grid.text(nm[i], x = x, y = y)
        } else if(i > j) {
            grid.circle(x = x, y = y, r = abs(cor_mat[i, j])/2 * min(unit.c(width, height)), 
                gp = gpar(fill = col_fun(cor_mat[i, j]), col = NA))
        } else {
            grid.text(sprintf("%.1f", cor_mat[i, j]), x, y, gp = gpar(fontsize = 10))
        }
    }, cluster_rows = FALSE, cluster_columns = FALSE,
    show_row_names = FALSE, show_column_names = FALSE)
```

As you may see in previous plot, when setting the non-standard parameter
`rect_gp = gpar(type = "none")`, the clustering is performed but nothing is
drawn on the heatmap body.

One last example is to visualize a [GO
game](https://en.wikipedia.org/wiki/Go_%28game%29). The input data takes
records of moves in the game.

```{r}
str = "B[cp];W[pq];B[dc];W[qd];B[eq];W[od];B[de];W[jc];B[qk];W[qn]
;B[qh];W[ck];B[ci];W[cn];B[hc];W[je];B[jq];W[df];B[ee];W[cf]
;B[ei];W[bc];B[ce];W[be];B[bd];W[cd];B[bf];W[ad];B[bg];W[cc]
;B[eb];W[db];B[ec];W[lq];B[nq];W[jp];B[iq];W[kq];B[pp];W[op]
;B[po];W[oq];B[rp];W[ql];B[oo];W[no];B[pl];W[pm];B[np];W[qq]
;B[om];W[ol];B[pk];W[qp];B[on];W[rm];B[mo];W[nr];B[rl];W[rk]
;B[qm];W[dp];B[dq];W[ql];B[or];W[mp];B[nn];W[mq];B[qm];W[bp]
;B[co];W[ql];B[no];W[pr];B[qm];W[dd];B[pn];W[ed];B[bo];W[eg]
;B[ef];W[dg];B[ge];W[gh];B[gf];W[gg];B[ek];W[ig];B[fd];W[en]
;B[bn];W[ip];B[dm];W[ff];B[cb];W[fe];B[hp];W[ho];B[hq];W[el]
;B[dl];W[fk];B[ej];W[fp];B[go];W[hn];B[fo];W[em];B[dn];W[eo]
;B[gp];W[ib];B[gc];W[pg];B[qg];W[ng];B[qc];W[re];B[pf];W[of]
;B[rc];W[ob];B[ph];W[qo];B[rn];W[mi];B[og];W[oe];B[qe];W[rd]
;B[rf];W[pd];B[gm];W[gl];B[fm];W[fl];B[lj];W[mj];B[lk];W[ro]
;B[hl];W[hk];B[ik];W[dk];B[bi];W[di];B[dj];W[dh];B[hj];W[gj]
;B[li];W[lh];B[kh];W[lg];B[jn];W[do];B[cl];W[ij];B[gk];W[bl]
;B[cm];W[hk];B[jk];W[lo];B[hi];W[hm];B[gk];W[bm];B[cn];W[hk]
;B[il];W[cq];B[bq];W[ii];B[sm];W[jo];B[kn];W[fq];B[ep];W[cj]
;B[bk];W[er];B[cr];W[gr];B[gk];W[fj];B[ko];W[kp];B[hr];W[jr]
;B[nh];W[mh];B[mk];W[bb];B[da];W[jh];B[ic];W[id];B[hb];W[jb]
;B[oj];W[fn];B[fs];W[fr];B[gs];W[es];B[hs];W[gn];B[kr];W[is]
;B[dr];W[fi];B[bj];W[hd];B[gd];W[ln];B[lm];W[oi];B[oh];W[ni]
;B[pi];W[ki];B[kj];W[ji];B[so];W[rq];B[if];W[jf];B[hh];W[hf]
;B[he];W[ie];B[hg];W[ba];B[ca];W[sp];B[im];W[sn];B[rm];W[pe]
;B[qf];W[if];B[hk];W[nj];B[nk];W[lr];B[mn];W[af];B[ag];W[ch]
;B[bh];W[lp];B[ia];W[ja];B[ha];W[sf];B[sg];W[se];B[eh];W[fh]
;B[in];W[ih];B[ae];W[so];B[af]"
```

We convert it into a matrix:

```{r}
str = gsub("\\n", "", str)
step = strsplit(str, ";")[[1]]
type = gsub("(B|W).*", "\\1", step)
row = gsub("(B|W)\\[(.).\\]", "\\2", step)
column = gsub("(B|W)\\[.(.)\\]", "\\2", step)

go_mat = matrix(nrow = 19, ncol = 19)
rownames(go_mat) = letters[1:19]
colnames(go_mat) = letters[1:19]
for(i in seq_along(row)) {
    go_mat[row[i], column[i]] = type[i]
}
go_mat[1:4, 1:4]
```

Black and white stones are put based on the values in the matrix:

```{r, fig.width = 6, fig.height = 6}
Heatmap(go_mat, name = "go", rect_gp = gpar(type = "none"),
    cell_fun = function(j, i, x, y, w, h, col) {
        grid.rect(x, y, w, h, gp = gpar(fill = "#dcb35c", col = NA))
        if(i == 1) {
            grid.segments(x, y-h*0.5, x, y)
        } else if(i == nrow(go_mat)) {
            grid.segments(x, y, x, y+h*0.5)
        } else {
            grid.segments(x, y-h*0.5, x, y+h*0.5)
        }
        if(j == 1) {
            grid.segments(x, y, x+w*0.5, y)        
        } else if(j == ncol(go_mat)) {
            grid.segments(x-w*0.5, y, x, y)
        } else {
            grid.segments(x-w*0.5, y, x+w*0.5, y)
        }

        if(i %in% c(4, 10, 16) & j %in% c(4, 10, 16)) {
            grid.points(x, y, pch = 16, size = unit(2, "mm"))
        }
                
        r = min(unit.c(w, h))*0.45
        if(is.na(go_mat[i, j])) {
        } else if(go_mat[i, j] == "W") {
            grid.circle(x, y, r, gp = gpar(fill = "white", col = "white"))
        } else if(go_mat[i, j] == "B") {
            grid.circle(x, y, r, gp = gpar(fill = "black", col = "black"))
        }
    },
    col = c("B" = "black", "W" = "white"),
    show_row_names = FALSE, show_column_names = FALSE,
    column_title = "One famous GO game",
    heatmap_legend_param = list(title = "Player", at = c("B", "W"), 
        labels = c("player1", "player2"), border = "black")
)
```

### layer_fun {#layer-fun}

`cell_fun` adds graphics cell by cell, while `layer_fun` adds graphics in a
block-wise manner. Similar as `cell_fun`, `layer_fun` also needs seven
arguments, but they are all in vector form (`layer_fun` can also have a eighth and
ninth arguments which is introduced later in this section):

```{r, eval = FALSE}
# code only for demonstration
Heatmap(..., layer_fun = function(j, i, x, y, w, h, fill) {...})
# or you can capitalize the arguments to mark they are vectors,
# the names of the argumetn do not matter
Heatmap(..., layer_fun = function(J, I, X, Y, W, H, F) {...})
```

``j`` and ``i`` still contain the column and row indices corresponding to the
original matrix, but since now `layer_fun` applies to a block of cells (or a
block of heatmap if the heatmap is split), ``j`` and ``i`` are vectors for all
the cells in the current heatmap slice. Similarlly, `x`, `y`, `w`, `h` and
`fill` are all vectors corresponding to all cells in the current heatmap
slice.

Since `j` and `i` now are vectors, to get corresponding values in the matrix, we
cannot use the form as `mat[j, i]` because it gives you a sub-matrix with
`length(i)` rows and `length(j)` columns. Instead we can use `pindex()`
function from **ComplexHeatmap** which is like pairwise indexing for a matrix.
See follow example:

```{r}
mfoo = matrix(1:9, nr = 3)
mfoo[1:2, c(1, 3)]
# but we actually want mfoo[1, 1] and mfoo[2, 3]
pindex(mfoo, 1:2, c(1, 3))
```

Next example shows the `layer_fun` version of adding text on heatmap. It's
basically the same as the `cell_fun` version.

```{r, fig.width = 5.2}
col_fun = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
Heatmap(small_mat, name = "mat", col = col_fun,
    layer_fun = function(j, i, x, y, width, height, fill) {
        # since grid.text can also be vectorized
        grid.text(sprintf("%.1f", pindex(small_mat, i, j)), x, y, gp = gpar(fontsize = 10))
})
```

And only add text to cells with positive values:

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat", col = col_fun, 
    layer_fun = function(j, i, x, y, width, height, fill) {
        v = pindex(small_mat, i, j)
        l = v > 0
        grid.text(sprintf("%.1f", v[l]), x[l], y[l], gp = gpar(fontsize = 10))
})
```

When the heatmap is split, `layer_fun` is applied in every slice.

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat", col = col_fun,
    row_km = 2, column_km = 2,
    layer_fun = function(j, i, x, y, width, height, fill) {
        v = pindex(small_mat, i, j)
        grid.text(sprintf("%.1f", v), x, y, gp = gpar(fontsize = 10))
        if(sum(v > 0)/length(v) > 0.75) {
            grid.rect(gp = gpar(lwd = 2, fill = "transparent"))
        }
})
```

`layer_fun` can also have two more arguments which are the index for the
current row slice and column slice. E.g. we want to add borders for the top
right and bottom left slices.

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat", col = col_fun,
    row_km = 2, column_km = 2,
    layer_fun = function(j, i, x, y, width, height, fill, slice_r, slice_c) {
        v = pindex(small_mat, i, j)
        grid.text(sprintf("%.1f", v), x, y, gp = gpar(fontsize = 10))
        if(slice_r != slice_c) {
            grid.rect(gp = gpar(lwd = 2, fill = "transparent"))
        }
})
```

The advantage of `layer_fun` is that it is not only fast to add graphics, but
also it provides more possibilities to customize the heatmap. Consider
following visualization: For each row in the heatmap, if values in the
neighbouring two columns have the same sign, we add a red line or a green line
depending on the sign of the two values. Since now graphics in a cell depend
on other cells, it is only possible to implement it via `layer_fun`. (Don't be
frightened by following code. They are explained after the code.)

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat", col = col_fun,
    row_km = 2, column_km = 2,
    layer_fun = function(j, i, x, y, w, h, fill) {
        # restore_matrix() is explained after this chunk of code
        ind_mat = restore_matrix(j, i, x, y)
        for(ir in seq_len(nrow(ind_mat))) {
            # start from the second column
            for(ic in seq_len(ncol(ind_mat))[-1]) {
                ind1 = ind_mat[ir, ic-1] # previous column
                ind2 = ind_mat[ir, ic]   # current column
                v1 = small_mat[i[ind1], j[ind1]]
                v2 = small_mat[i[ind2], j[ind2]]
                if(v1 * v2 > 0) { # if they have the same sign
                    col = ifelse(v1 > 0, "darkred", "darkgreen")
                    grid.segments(x[ind1], y[ind1], x[ind2], y[ind2],
                        gp = gpar(col = col, lwd = 2))
                    grid.points(x[c(ind1, ind2)], y[c(ind1, ind2)], 
                        pch = 16, gp = gpar(col = col), size = unit(4, "mm"))
                }
            }
        }
    }
)
```

The values that are sent to `layer_fun` are all vectors (for the vectorization
of the **grid** graphic functions), however, the heatmap slice where
`layer_fun` is applied to, is still represented by a matrix, thus, it would be
very convinient if all the arguments in `layer_fun` can be converted to the
sub-matrix for the current slice. Here, as shown in above example,
`restore_matrix()` does the job. `restore_matrix()` directly accepts the first
four argument in `layer_fun` and returns an index matrix, where rows and
columns correspond to the rows and columns in the current slice, from top to
bottom and from left to right. The values in the matrix are the natural order
of e.g. vector `j` in current slice.

If you run following code:

```{r, eval = FALSE}
Heatmap(small_mat, name = "mat", col = col_fun,
    row_km = 2, column_km = 2,
    layer_fun = function(j, i, x, y, w, h, fill) {
        ind_mat = restore_matrix(j, i, x, y)
        print(ind_mat)
    }
)
```

The first output which is for the top-left slice:

```
     [,1] [,2] [,3] [,4] [,5]
[1,]    1    4    7   10   13
[2,]    2    5    8   11   14
[3,]    3    6    9   12   15
```

As you see, this is a three-row and five-column index matrix where the first
row corresponds to the top row in the slice. The values in the matrix
correspond to the natural index (i.e. 1, 2, ...) in `j`, `i`, `x`, `y`,
... in `layer_fun`. Now, if we want to add values on the second column in the
top-left slice, the code which is put inside `layer_fun` would look like:

```{r, eval = FALSE}
for(ind in ind_mat[, 2]) {
    grid.text(small_mat[i[ind], j[ind]], x[ind], y[ind], ...)
}
```

Now it is easier to understand the second example: we want to add points to
the second row and the third column in every slice:

```{r, fig.width = 5.2}
Heatmap(small_mat, name = "mat", col = col_fun,
    row_km = 2, column_km = 2,
    layer_fun = function(j, i, x, y, w, h, fill) {
        ind_mat = restore_matrix(j, i, x, y)
        ind = unique(c(ind_mat[2, ], ind_mat[, 3]))
        grid.points(x[ind], y[ind], pch = 16, size = unit(4, "mm"))
    }
)
```

## Size of the heatmap {#size-of-the-heatmap}

`width`, `heatmap_width`, `height` and `heatmap_height` control the size of
the heatmap. By default, all heatmap components have fixed width or height,
e.g. the width of row dendrogram is `1cm`. The width or the height of the
heatmap body fill the rest area of the final plotting region, which means, if
you draw it in an interactive graphic window and you change the size of the
window by draging it, the size of the heatmap body is automatically adjusted.

`heatmap_width` and `heatmap_height` control **the width/height of the complete
heatmap including all heatmap components (excluding the legends)** while `width`
and `height` only **control the width/height of the heamtap body**. All these four
arguments can be set as absolute units.

```{r}
Heatmap(mat, name = "mat", width = unit(8, "cm"), height = unit(8, "cm"))
Heatmap(mat, name = "mat", heatmap_width = unit(8, "cm"), heatmap_height = unit(8, "cm"))
```

These four arguments are more important when adjust the size in a list of
heatmaps (see Section
\@ref(size-of-heatmaps)).

When the size of the heatmap is set as absolute units, it is possible that the
size of the figure is larger than the size of the heatmap, which gives blank
areas around the heatmap. The size of the heatmap can be retrieved by `ComplexHeatmap:::width()`
and `ComplexHeatmap:::height()` functions.

```{r, echo = 2:10}
pdf(NULL)
ht = Heatmap(mat, name = "mat", width = unit(8, "cm"), height = unit(8, "cm"))
ht = draw(ht) # you must call draw() to reassign the heatmap variable
ComplexHeatmap:::width(ht)
ComplexHeatmap:::height(ht)

ht = Heatmap(mat, name = "mat", heatmap_width = unit(8, "cm"), 
    heatmap_height = unit(8, "cm"))
ht = draw(ht)
ComplexHeatmap:::width(ht)
ComplexHeatmap:::height(ht)
invisible(dev.off())
```

Check this blog post ([Set cell width/height in the heatmap](https://jokergoo.github.io/2020/05/11/set-cell-width/height-in-the-heatmap/)) if you want to save the heatmap in a figure file which
has exactly the same size as the heatmap.

## Plot the heatmap {#plot-the-heatmap}

`Heatmap()` function actually is only a constructor, which means it only puts
all the data and configurations into the object in the `Heatmap` class. The
clustering will only be performed when the `draw()` method is called. Under
interactive mode (e.g. the interactive R terminal where you can type your R
code line by line), directly calling `Heatmap()` without returning to any
object prints the object and the print method (or the S4 `show()` method) for
the `Heatmap` class object calls `draw()` internally. So if you type
`Heatmap(...)` in your R terminal, it looks like it is a plotting function
like `plot()`, you need to be aware of that it is actually not true and in the
following cases you might see nothing plotted.

- you put `Heatmap(...)` inside a function,
- you put `Heatmap(...)` in a code chunk like `for` or `if-else`
- you put `Heatmap(...)` in an Rscript and you run it under command line.

The reason is in above three cases, the `show()` method WILL NOT be called and
thus `draw()` method is not executed either. So, to make the plot, you need to
call `draw()` explicitly: `draw(Heatmap(...))` or:

```{r, eval = FALSE}
# code only for demonstration
ht = Heatmap(...)
draw(ht)
```

The `draw()` function actually is applied to a list of heatmaps in
`HeatmapList` class. The `draw()` method for the single `Heatmap` class
constructs a `HeatmapList` with only one heatmap and call `draw()` method of
the `HeatmapList` class. The `draw()` function accpets a lot of more arguments
which e.g. controls the legends. It will be discussed in Chapter
\@ref(a-list-of-heatmaps).

```{r, eval = FALSE}
draw(ht, heatmap_legend_side, padding, ...)
```

## Get orders and dendrograms {#get-orders-and-dendrograms-from-heatmap}

The row/column orders of the heatmap can be obtained by
`row_order()`/`column_order()` functions. You can directly apply to the
heatmap object returned by `Heatmap()` or to the object returned by `draw()`.
In following, we take `row_order()` as example.

```{r, echo = 2:6}
pdf(NULL)
small_mat = mat[1:9, 1:9]
ht1 = Heatmap(small_mat)
row_order(ht1)
ht2 = draw(ht1)
row_order(ht2)
invisible(dev.off())
```

As explained in previous section, `Heatmap()` function does not perform
clustering, thus, when directly apply `row_order()` on `ht1`, clustering will
be first performed. Later when making the heatmap by `draw(ht1)`, the
clustering will be applied again. This might be a problem that if you set
k-means clustering in the heatmap. Since the clustering is applied twice,
k-means might give you different clusterings, which means, you might have
different results from `row_order()` and you might have different heatmap.

In following chunk of code, `o1`, `o2` and `o3` might be different because
each time, k-means clustering is performed.

```{r, eval = FALSE}
# code only for demonstration
ht1 = Heatmap(small_mat, row_km = 2)
o1 = row_order(ht1)
o2 = row_order(ht1)
ht2 = draw(ht1)
o3 = row_order(ht2)
o4 = row_order(ht2)
```

`draw()` function returns the heatmap (or more precisely, the heatmap list)
which has already been reordered, and applying `row_order()` just extracts the
row order from the object, which ensures the row order is exactly the same as
the one shown in the heatmap. In above code, `o3` is always identical to `o4`.

So, the preferable way to get row/column orders is as follows.

```{r, eval = FALSE}
# code only for demonstration
ht = Heatmap(small_mat)
ht = draw(ht)
row_order(ht)
column_order(ht)
```

If rows/columns are split, row order or column order will be a list.

```{r, echo = 2:5}
pdf(NULL)
ht = Heatmap(small_mat, row_km = 2, column_km = 3)
ht = draw(ht)
row_order(ht)
column_order(ht)
invisible(dev.off())
```

Similarly, the `row_dend()`/`column_dend()` functions return the dendrograms.
It returns a single dendrogram or a list of dendrograms depending on whether
the heatmap is split.

```{r, echo = 2:5}
pdf(NULL)
ht = Heatmap(small_mat, row_km = 2)
ht = draw(ht)
row_dend(ht)
column_dend(ht)
invisible(dev.off())
```

`row_order()`, `column_order()`, `row_dend()` and `column_dend()` also work
for a list of heatmaps, it will be introduced in Section
\@ref(get-orders-and-dendrograms-from-a-list-of-heatmaps).

## Subset a heatmap {#subset-a-heatmap}

Since heatmap is a representation of a matrix, there is also a subset method
for the `Heatmap` class.

```{r, fig.width = 5.2}
ht = Heatmap(mat, name = "mat")
dim(ht)
ht[1:10, 1:10]
```

The annotations are subsetted accordingly as well.

```{r}
ht = Heatmap(mat, name = "mat", row_km = 2, column_km = 3,
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    top_annotation = HeatmapAnnotation(foo1 = 1:24, bar1 = anno_points(runif(24))),
    right_annotation = rowAnnotation(foo2 = 18:1, bar2 = anno_barplot(runif(18)))
)
ht[1:9*2 - 1, 1:12*2] # odd rows, even columns
```

The heatmap components are subsetted if they are vector-like. Some
configurations in the heatmap keep the same when subsetting, e.g. if `row_km`
is set in the original heatmap, the configuration of k-means is kept and it is
performed in the sub-heatmap. So in following example, k-means clustering is
only performed when making heatmap for `ht2`.

```{r, fig.width = 5.2}
ht = Heatmap(mat, name = "mat", row_km = 2)
ht2 = ht[1:10, 1:10]
ht2
```

**The implementation of subsetting heatmaps is very experimental.** It is not
always working, e.g. if `cell_fun` is defined and uses an external matrix, or
clustering objects are assigned to `cluster_rows` or `cluster_columns`.

There are also subset methods for the `HeatmapAnnotation` class (Section
\@ref(heatmap-annotation-utility-function)) and the `HeatmapList` class (Section \@ref(subset-heatmap-list)),
but both are very experimental as well.


