# UpSet plot {#upset-plot}

```{r, echo = FALSE, message = FALSE}
library(circlize)
library(GenomicRanges)
```

[UpSet plot](https://caleydo.org/tools/upset/) provides an efficient way to
visualize intersections of multiple sets compared to the traditional
approaches, i.e. the Venn Diagram. It is implemented in the [UpSetR
package](https://github.com/hms-dbmi/UpSetR) in R. Here we re-implemented
UpSet plots with the **ComplexHeatmap** package with some improvements.

## Input data {#input-data}

To represent multiple sets, the variable can be represented as:

1. A list of sets where each set is a vector, e.g.:

```{r, eval = FALSE}
list(set1 = c("a", "b", "c"),
     set2 = c("b", "c", "d", "e"),
     ...)
```

2. A binary matrix/data frame where rows are elements and columns are sets,
   e.g.:

```
  set1 set2 set3
h    1    1    1
t    1    0    1
j    1    0    0
u    1    0    1
w    1    0    0
...
```

E.g., for row `t`, it means, `t` is in set **set1**, not in set **set2**, and
in set **set3**. Note the matrix is also valid if it is a logical matrix.

If the variable is a data frame, the binary columns (only contain 0 and 1) and
the logical columns are only used.

Both formats can be used for making UpSet plots, users can still use
`list_to_matrix()` to convert from list to the binary matrix.

```{r}
lt = list(set1 = c("a", "b", "c"),
          set2 = c("b", "c", "d", "e"))
list_to_matrix(lt)
```

You can also set the universal set in `list_to_matrix()`:

```{r}
list_to_matrix(lt, universal = letters[1:10])
```

3. The set can be genomic intervals, then it can only be represented as a list
   of `GRanges`/`IRanges` objects.

```{r, eval = FALSE}
list(set1 = GRanges(...),
	 set2 = GRanges(...),
	 ...)
```

## Mode {#upset-mode}

E.g. for three sets (**A**, **B**, **C**), all combinations of selecting
elements in the set or not in the set are as following:

```
A B C
1 1 1
1 1 0
1 0 1
0 1 1
1 0 0
0 1 0
0 0 1
```

A value of 1 means to select that set and 0 means not to select that set.
E.g., "1 1 0" means to select set A, B while not set C. Note there is no "0 0
0", because the background set is not of interest here. In following part of
this section, we refer **A**, **B** and **C** as **sets** and each combination
as **combination set**. The whole binary matrix is called **combination
matrix**.

The UpSet plot visualizes the **size** of each combination set. With the
binary code of each combination set, next we need to define how to calculate
the size of that combination set. There are three modes:

1. `distinct` mode: 1 means in that set and 0 means not in that set, then `1 1
   0` means a set of elements both in set **A** and **B**, while not in **C**
   (`setdiff(intersect(A, B), C)`). Under this mode, the seven combination
   sets are the seven partitions in the Venn diagram and they are mutually
   exclusive.

2. `intersect` mode: 1 means in that set and 0 is not taken into account,
   then, `1 1 0` means a set of elements in set **A** and **B**, and they can
   also in **C** or not in **C** (`intersect(A, B)`). Under this mode, the
   seven combination sets can overlap.

3. `union mode`: 1 means in that set and 0 is not taken into account. When
   there are multiple 1, the relationship is __OR__. Then, `1 1 0` means a set
   of elements in set **A** or **B**, and they can also in **C** or not in
   **C** (`union(A, B)`). Under this mode, the seven combination sets can
   overlap.

The three modes are illustrated in following figure:

```{r, fig.width = 6*0.8, fig.height = 8*0.8, echo = FALSE}
source("upset_mode.R")
```

## Make the combination matrix {#make-the-combination-matrix}

The function `make_comb_mat()` generates the combination matrix as well as
calculates the size of the sets and the combination sets. The input can be one
single variable or name-value pairs:

```{r}
set.seed(123)
lt = list(a = sample(letters, 5),
	      b = sample(letters, 10),
	      c = sample(letters, 15))
m1 = make_comb_mat(lt)
m1
m2 = make_comb_mat(a = lt$a, b = lt$b, c = lt$c)
m3 = make_comb_mat(list_to_matrix(lt))
```

`m1`, `m2` and `m3` are identical.

The mode is controlled by the `mode` argument:

```{r}
m1 = make_comb_mat(lt) # the default mode is `distinct`
m2 = make_comb_mat(lt, mode = "intersect")
m3 = make_comb_mat(lt, mode = "union")
```

The UpSet plots under different modes will be demonstrated in later sections.

When there are too many sets, the sets can be pre-filtered by the set sizes.
The `min_set_size` and `top_n_sets` are for this purpose. `min_set_size`
controls the minimal size for the sets and `top_n_sets` controls the number of
top sets with largest sizes.

```{r}
m1 = make_comb_mat(lt, min_set_size = 4)
m2 = make_comb_mat(lt, top_n_sets = 2)
```

The subsetting of the sets affects the calculation of the sizes of the
combination sets, that is why it needs to be controlled at the combination
matrix generation step. The subsetting of combination sets can be directly
performed by subsetting the matrix:

```{r}
m = make_comb_mat(lt)
m[1:4]
```

`make_comb_mat()` also allows to specify the universal set so that the complement
set which contains elements not belonging to any set is also considered.

```{r}
m = make_comb_mat(lt, universal_set = letters)
m
```

The universal set can be smaller than the union of all sets, then for each
set, only the intersection to universal set is considered.

```{r}
m = make_comb_mat(lt, universal_set = letters[1:10])
m
```

If you already know the size of the complement size, you can directly set 
`complement_size` argument.

```{r}
m = make_comb_mat(lt, complement_size = 5)
m
```

When the input is matrix and it contains elements that do not belong to any of
the set, these elements are treated as complement set.

```{r}
x = list_to_matrix(lt, universal_set = letters)
m = make_comb_mat(x)
m
```

The universal set also works for sets as genomic regions.

## Utility functions {#upset-utility-functions}

`make_comb_mat()` returns a matrix, also in `comb_mat` class. There are some
utility functions that can be applied to this `comb_mat` object:

- `set_name()`: The set names.
- `comb_name()`: The combination set names. The names of the combination sets
  are formatted as a string of binary bits. E.g. for three sets of **A**,
  **B**, **C**, the combination set with name "101" corresponds to selecting set
  **A**, not selecting set **B** and selecting set **C**.
- `set_size()`: The set sizes.
- `comb_size()`: The combination set sizes.
- `comb_degree()`: The degree for a combination set is the number of sets that
  are selected.
- `t()`: Transpose the combination matrix. By default `make_comb_mat()`
  generates a matrix where sets are on rows and combination sets are on
  columns, and so are they on the UpSet plots. By transposing the combination
  matrix, the position of sets and combination sets can be swtiched on the
  UpSet plot.
- `extract_comb()`: Extract the elements in a specified combination set. The
  usage will be explained later.

Quick examples are:

```{r}
m = make_comb_mat(lt)
set_name(m)
comb_name(m)
set_size(m)
comb_size(m)
comb_degree(m)
t(m)
```

For using `extract_comb()`, the valid combination set name should be from
`comb_name()`. Note the elements in the combination sets depends on the "mode"
set in `make_comb_mat()`.

```{r}
extract_comb(m, "101")
```

Next we demonstrate a second example, where the sets are genomic regions.
**When the sets are genomic regions, the size is calculated as the sum of the
width of regions in each set (or in other words, the total number of base
pairs).**

```{r}
library(circlize)
library(GenomicRanges)
lt2 = lapply(1:4, function(i) generateRandomBed())
lt2 = lapply(lt2, function(df) GRanges(seqnames = df[, 1], 
	ranges = IRanges(df[, 2], df[, 3])))
names(lt2) = letters[1:4]
m = make_comb_mat(lt2)
set_size(m)
comb_size(m)
```

And now `extract_comb()` returns genomic regions that are in the corresponding combination set.

```{r}
extract_comb(m, "1010")
```

With `comb_size()` and `comb_degree()`, we can filter the combination matrix as:

```{r}
m = make_comb_mat(lt)
# combination set size >= 4
m[comb_size(m) >= 4]
# combination set degree == 2
m[comb_degree(m) == 2]
```

For the complement set, the name for this special combination set is only
composed of zeros.

```{r}
m2 = make_comb_mat(lt, universal_set = letters)
comb_name(m2) # see the first element
comb_degree(m2)
```

If `universal_set` was set in `make_comb_mat()`, `extract_comb()` can be
applied to the complement set.

```{r}
m2 = make_comb_mat(lt, universal_set = letters)
extract_comb(m2, "000")
m2 = make_comb_mat(lt, universal_set = letters[1:10])
extract_comb(m2, "000")
```

When `universal_set` was set, `extract_comb()` also works for genomic region sets.

## Make the plot {#upset-making-the-plot}

Making the UpSet plot is very straightforward that users just send the
combination matrix to `UpSet()` function:

```{r, fig.width = 5, fig.height = 3}
UpSet(m)
```

By default the sets are ordered by the size and the combination sets are
ordered by the degree (number of sets that are selected).

The order is controlled by `set_order` and `comb_order`:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, set_order = c("a", "b", "c"), comb_order = order(comb_size(m)))
```

Color of dots, size of dots and line width of the segments are controlled by
`pt_size`, `comb_col` and `lwd`. `comb_col` should be a vector corresponding
to the combination sets. In following code, since `comb_degree(m)` returns a
vector of integers, we just use it as index for the color vector.

```{r, fig.width = 5, fig.height = 3}
UpSet(m, pt_size = unit(5, "mm"), lwd = 3,
	comb_col = c("red", "blue", "black")[comb_degree(m)])
```

Colors for the background (the rectangles and the dots representing the set is
not selected) are controlled by `bg_col`, `bg_pt_col`. The length of `bg_col`
can have length of one or two.

```{r, fig.width = 5, fig.height = 3}
UpSet(m, comb_col = "#0000FF", bg_col = "#F0F0FF", bg_pt_col = "#CCCCFF")
UpSet(m, comb_col = "#0000FF", bg_col = c("#F0F0FF", "#FFF0F0"), bg_pt_col = "#CCCCFF")
```

Transposing the combination matrix swtiches the sets to columns and
combination sets to rows.

```{r, fig.width = 3, fig.height = 5}
UpSet(t(m))
```

As we have introduced, if do subsetting on the combination sets, the subset of
the matrix can be visualized as well:

```{r, eval = FALSE}
UpSet(m[comb_size(m) >= 4])
UpSet(m[comb_degree(m) == 2])
```

```{r, echo = FALSE, fig.height = 3, fig.width = 7}
grid.newpage()
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(UpSet(m[comb_size(m) >= 4], column_title = "comb_size(m) >= 4"), newpage = FALSE)
popViewport()
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(UpSet(m[comb_degree(m) == 2], column_title = "comb_degree(m) == 2"), newpage = FALSE)
popViewport()
```

Following compares the different mode in `make_comb_mat()`:

```{r, eval = FALSE}
m1 = make_comb_mat(lt) # the default mode is `distinct`
m2 = make_comb_mat(lt, mode = "intersect")
m3 = make_comb_mat(lt, mode = "union")
UpSet(m1)
UpSet(m2)
UpSet(m3)
```

```{r, echo = FALSE, fig.height = 8, fig.width = 5}
m1 = make_comb_mat(lt) # the default mode is `distinct`
m2 = make_comb_mat(lt, mode = "intersect")
m3 = make_comb_mat(lt, mode = "union")
grid.newpage()
pushViewport(viewport(y = 1, height = 1/3, just = "top"))
draw(UpSet(m1, column_title = "mode = 'distinct'"), newpage = FALSE)
popViewport()
pushViewport(viewport(y = 2/3, height = 1/3, just = "top"))
draw(UpSet(m2, column_title = "mode = 'intersect'"), newpage = FALSE)
popViewport()
pushViewport(viewport(y = 1/3, height = 1/3, just = "top"))
draw(UpSet(m3, column_title = "mode = 'union'"), newpage = FALSE)
popViewport()
```

For the plot containing complement set, there is one additional column showing
this complement set does not overlap to any of the sets (all dots are in grey).

```{r, fig.width = 5, fig.height = 3}
m2 = make_comb_mat(lt, universal_set = letters)
UpSet(m2)
```

If you already know the size for the complement set, you can directly assign it 
by `complement_size` argument in `make_comb_mat()`.

```{r, fig.width = 5, fig.height = 3}
m2 = make_comb_mat(lt, complement_size = 10)
UpSet(m2)
```

For the case where the universal set is smaller than the union of all sets.

```{r, fig.width = 5, fig.height = 3}
m2 = make_comb_mat(lt, universal_set = letters[1:10])
UpSet(m2)
```

By default the empty combination sets are removed from the plot, but they can be 
kept by setting `remove_empty_comb_set = FALSE` in `make_comb_mat()`.

```{r, fig.width = 5, fig.height = 3}
m2 = make_comb_mat(lt, universal_set = letters[1:10], remove_empty_comb_set = FALSE)
UpSet(m2)
```

There are some cases that you may have complement set but you don't want to show it,
especially when the input for `make_comb_mat()` is a matrix which already contains 
complement set.

```{r, fig.width = 5, fig.height = 3}
x = list_to_matrix(lt, universal_set = letters)
m2 = make_comb_mat(x)
UpSet(m2)
m2 = make_comb_mat(x, remove_complement_set = TRUE)
UpSet(m2)
```

Setting `remove_complement_set = TRUE` is identical to:

```{r, eval = FALSE}
m2 = make_comb_mat(x)
m2 = m2[comb_degree(m2) > 0]
```

## UpSet plots as heatmaps {#upset-plots-as-heatmaps}

In the UpSet plot, the major component is the combination matrix, and on the
two sides are the barplots representing the size of sets and the combination
sets, thus, it is quite straightforward to implement it as a "heatmap" where
the heatmap is self-defined with dots and segments, and the two barplots are
two barplot annotations constructed by `anno_barplot()`.

The default top annotation is:

```{r, eval = FALSE}
HeatmapAnnotation("Intersection\nsize" = anno_barplot(comb_size(m), 
		border = FALSE, gp = gpar(fill = "black"), height = unit(3, "cm")), 
	annotation_name_side = "left", annotation_name_rot = 0)
```

This top annotation is wrapped in `upset_top_annotation()` which only contais
the upset top barplot annotation. Most of the arguments in
`upset_top_annotation()` directly goes to the `anno_barplot()`, e.g. to set
the colors of bars:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, top_annotation = upset_top_annotation(m, 
	gp = gpar(col = comb_degree(m))))
```

To control the data range and axis:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, top_annotation = upset_top_annotation(m, 
	ylim = c(0, 15),
	bar_width = 1,
	axis_param = list(side = "right", at = c(0, 5, 10, 15),
		labels = c("zero", "five", "ten", "fifteen"))))
```

To control the annotation name:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, top_annotation = upset_top_annotation(m, 
	annotation_name_rot = 90,
	annotation_name_side = "right",
	axis_param = list(side = "right")))
```

The settings are very similar for the right annotation:


```{r, fig.width = 5, fig.height = 3}
UpSet(m, right_annotation = upset_right_annotation(m, 
	ylim = c(0, 30),
	gp = gpar(fill = "green"),
	annotation_name_side = "top",
	axis_param = list(side = "top")))
```

`upset_top_annotation()` and `upset_right_annotation()` can automatically
recognize whether sets are on rows or columns.


`upset_top_annotation()` and `upset_right_annotation()` only contain one
barplot annotation. If users want to add more annotations, they need to
manually construct a `HeatmapAnnotation` object with multiple annotations.

To add more annotations on top:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, top_annotation = HeatmapAnnotation(
	degree = as.character(comb_degree(m)),
	"Intersection\nsize" = anno_barplot(comb_size(m), 
		border = FALSE, 
		gp = gpar(fill = "black"), 
		height = unit(2, "cm")
	), 
	annotation_name_side = "left", 
	annotation_name_rot = 0))
```

To add more annotation on the right:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, right_annotation = rowAnnotation(
	"Set size" = anno_barplot(set_size(m), 
		border = FALSE, 
		gp = gpar(fill = "black"), 
		width = unit(2, "cm")
	),
	group = c("group1", "group1", "group2")))
```

To move the right annotation to the left of the combination matrix:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, left_annotation = rowAnnotation(
	"Set size" = anno_barplot(set_size(m), 
		border = FALSE, 
		gp = gpar(fill = "black"), 
		width = unit(2, "cm")
	)), right_annotation = NULL)
```

To reverse the axis of the left annotation:

```{r, fig.width = 5, fig.height = 3}
UpSet(m, left_annotation = rowAnnotation(
	"Set size" = anno_barplot(set_size(m), 
		axis_param = list(direction = "reverse"),
		border = FALSE, 
		gp = gpar(fill = "black"), 
		width = unit(2, "cm")
	)), right_annotation = NULL,
	row_names_side = "right")
```

The object returned by `UpSet()` is actually a `Heatmap` class object, thus,
you can add to other heatmaps and annotations by `+` or `%v%`.

```{r, fig.width = 6, fig.height = 3}
ht = UpSet(m)
class(ht)
ht + Heatmap(1:3, name = "foo", width = unit(5, "mm")) + 
	rowAnnotation(bar = anno_points(1:3))
```

```{r, fig.width = 5, fig.height = 3}
ht %v% Heatmap(rbind(1:7), name = "foo", row_names_side = "left", 
		height = unit(5, "mm")) %v% 
	HeatmapAnnotation(bar = anno_points(1:7),
		annotation_name_side = "left")
```

Add multiple UpSet plots:

```{r, fig.width = 5, fig.height = 7}
m1 = make_comb_mat(lt, mode = "distinct")
m2 = make_comb_mat(lt, mode = "intersect")
m3 = make_comb_mat(lt, mode = "union")
UpSet(m1, row_title = "distinct mode") %v%
	UpSet(m2, row_title = "intersect mode") %v%
	UpSet(m3, row_title = "union mode")
```

Or first transpose all the combination matrices and add them horizontally:

```{r, fig.height = 5, fig.width = 7}
m1 = make_comb_mat(lt, mode = "distinct")
m2 = make_comb_mat(lt, mode = "intersect")
m3 = make_comb_mat(lt, mode = "union")
UpSet(t(m1), column_title = "distinct mode") +
	UpSet(t(m2), column_title = "intersect mode") +
	UpSet(t(m3), column_title = "union mode")
```

The three combination matrices are actually the same and plotting them three
times is redundant. With the functionality in **ComplexHeatmap** package, we
can use other types of annotations.

```{r, fig.width = 5, fig.height = 3}
# it is the same with using m1, m2 or m3
ht = UpSet(m1, top_annotation = HeatmapAnnotation(size = anno_lines(
		cbind(comb_size(m1), comb_size(m2), comb_size(m3)),
		gp = gpar(col = 2:4), height = unit(3, "cm")
)))
# you need to manually construct a legend
draw(ht, annotation_legend_list = list(Legend(
		title = "mode",
		type = "lines",
		labels = c("distinct", "intersect", "union"),
		legend_gp = gpar(col = 2:4)
	))
)
```

## Example with the movies dataset

[UpsetR package](https://github.com/hms-dbmi/UpSetR) also provides a `movies`
dataset, which contains 17 genres for 3883 movies. First load the dataset.

```{r}
movies = read.csv(system.file("extdata", "movies.csv", package = "UpSetR"), 
    header = TRUE, sep = ";")
head(movies)
```

To make a same UpSet plot as in [this vignette](https://cran.r-project.org/web/packages/UpSetR/vignettes/basic.usage.html#example-2-choosing-the-top-largest-sets-and-plot-formatting):

```{r, fig.width = 10, fig.height = 4}
m = make_comb_mat(movies, top_n_sets = 6, remove_complement_set = TRUE)
UpSet(m)
```

Following code makes it look more similar as the orignal plot. The code is a
little bit long, but most of the code mainly customize the annotations and
row/column orders.

```{r, fig.width = 10, fig.height = 4}
m = make_comb_mat(movies, top_n_sets = 6, remove_complement_set = TRUE)
ss = set_size(m)
UpSet(m, 
	set_order = order(set_size(m)),
	comb_order = order(comb_degree(m), -comb_size(m)),
	top_annotation = HeatmapAnnotation(
		"Genre Intersections" = anno_barplot(comb_size(m), 
			border = FALSE, 
			gp = gpar(fill = "black"), 
			height = unit(4, "cm")
		), 
		annotation_name_side = "left", 
		annotation_name_rot = 90),
	left_annotation = rowAnnotation(
		"Movies Per Genre" = anno_barplot(-ss, 
			baseline = 0,
			axis_param = list(
				at = c(0, -500, -1000, -1500),
				labels = c(0, 500, 1000, 1500),
				labels_rot = 0),
			border = FALSE, 
			gp = gpar(fill = "black"), 
			width = unit(4, "cm")
		),
		set_name = anno_text(set_name(m), 
			location = 0.5, 
			just = "center",
			width = max_text_width(set_name(m)) + unit(4, "mm"))
	), 
	right_annotation = NULL,
	show_row_names = FALSE)
```

In `movies` dataset, there is also one column `AvgRatinh` which gives the
rating of each movie, we next split all the movies into five groups based on
the ratings.

```{r}
genre = c("Action", "Romance", "Horror", "Children", "SciFi", "Documentary")
rating = cut(movies$AvgRating, c(0, 1, 2, 3, 4, 5))
m_list = tapply(seq_len(nrow(movies)), rating, function(ind) {
	make_comb_mat(movies[ind, genre, drop = FALSE], remove_complement_set = TRUE)
})
```

The combination matrices in `m_list` might have different combination sets:

```{r}
sapply(m_list, comb_size)
```

To compare between multiple groups with UpSet plots, we need to normalize all
the matrices to make them have same sets and same combination sets.
`normalize_comb_mat()` basically adds zero to the new combination sets which
were not there before.

```{r}
m_list = normalize_comb_mat(m_list)
sapply(m_list, comb_size)
```

We calculate the range for the two barplots:

```{r}
max_set_size = max(sapply(m_list, set_size))
max_comb_size = max(sapply(m_list, comb_size))
```

And finally we add the five UpSet plots vertically:

```{r, fig.width = 7, fig.height = 12}
ht_list = NULL
for(i in seq_along(m_list)) {
	ht_list = ht_list %v%
		UpSet(m_list[[i]], row_title = paste0("rating in", names(m_list)[i]),
			set_order = NULL, comb_order = NULL,
			top_annotation = upset_top_annotation(m_list[[i]], ylim = c(0, max_comb_size)),
			right_annotation = upset_right_annotation(m_list[[i]], ylim = c(0, max_set_size)))
}
ht_list
```

After comparing the five UpSet plots, we can see most of the movies are rated
between 2 and 4. Horror movies tend to have lower ratings and romance moves
tend to have higher ratings.

Instead of directly comparing the size of the combination sets, we can also
compare the relative fraction to the full sets. In following code, we remove
the group of `c(0, 1]` because the number of movies are too few there.

```{r, fig.width = 7, fig.height = 9}
m_list = m_list[-1]
max_set_size = max(sapply(m_list, set_size))
rel_comb_size = sapply(m_list, function(m) {
	s = comb_size(m)
	# because the combination matrix is generated under "distinct" mode
	# the sum of `s` is the size of the full set
	s/sum(s)
})
ht_list = NULL
for(i in seq_along(m_list)) {
	ht_list = ht_list %v%
		UpSet(m_list[[i]], row_title = paste0("rating in", names(m_list)[i]),
			set_order = NULL, comb_order = NULL,
			top_annotation = HeatmapAnnotation(
				"Relative\nfraction" = anno_barplot(
					rel_comb_size[, i],
					ylim = c(0, 0.5),
					gp = gpar(fill = "black"),
					border = FALSE,
					height = unit(2, "cm"),
				), 
				annotation_name_side = "left",
				annotation_name_rot = 0),
			right_annotation = upset_right_annotation(m_list[[i]], 
				ylim = c(0, max_set_size))
		)
}
ht_list
```

Now the trend is more clear that horror movies are rated low and documentaries
are rated high.

Next we split the movies by years:

```{r, fig.width = 14, fig.height = 10}
year = floor(movies$ReleaseDate/10)*10
m_list = tapply(seq_len(nrow(movies)), year, function(ind) {
	make_comb_mat(movies[ind, genre, drop = FALSE], remove_complement_set = TRUE)
})
m_list = normalize_comb_mat(m_list)
max_set_size = max(sapply(m_list, set_size))
max_comb_size = max(sapply(m_list, comb_size))
ht_list1 = NULL
for(i in 1:5) {
	ht_list1 = ht_list1 %v%
		UpSet(m_list[[i]], row_title = paste0(names(m_list)[i], "s"),
			set_order = NULL, comb_order = NULL,
			top_annotation = upset_top_annotation(m_list[[i]], ylim = c(0, max_comb_size),
				height = unit(2, "cm")),
			right_annotation = upset_right_annotation(m_list[[i]], ylim = c(0, max_set_size)))
}

ht_list2 = NULL
for(i in 6:10) {
	ht_list2 = ht_list2 %v%
		UpSet(m_list[[i]], row_title = paste0(names(m_list)[i], "s"),
			set_order = NULL, comb_order = NULL,
			top_annotation = upset_top_annotation(m_list[[i]], ylim = c(0, max_comb_size),
				height = unit(2, "cm")),
			right_annotation = upset_right_annotation(m_list[[i]], ylim = c(0, max_set_size)))
}
grid.newpage()
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(ht_list1, newpage = FALSE)
popViewport()
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(ht_list2, newpage = FALSE)
popViewport()
```

Now we can see most of the movies were produces in 1990s and the two major
genres are actions and romance.

Similarly, if we change the top annotation to the relative fraction to the
full sets (code not shown):

```{r, fig.width = 14, fig.height = 10, echo = FALSE}
max_set_size = max(sapply(m_list, set_size))
rel_comb_size = sapply(m_list, function(m) {
	s = comb_size(m)
	s/sum(s)
})
ht_list1 = NULL
for(i in 1:5) {
	ht_list1 = ht_list1 %v%
		UpSet(m_list[[i]], row_title = paste0(names(m_list)[i], "s"),
			set_order = NULL, comb_order = NULL,
			top_annotation = HeatmapAnnotation(
				"Relative\nfraction" = anno_barplot(
					rel_comb_size[, i],
					ylim = c(0, 0.5),
					gp = gpar(fill = "black"),
					border = FALSE,
					height = unit(2, "cm"),
				), 
				annotation_name_side = "left",
				annotation_name_rot = 0),
			right_annotation = upset_right_annotation(m_list[[i]], 
				ylim = c(0, max_set_size))
		)
}

ht_list2 = NULL
for(i in 6:10) {
	ht_list2 = ht_list2 %v%
		UpSet(m_list[[i]], row_title = paste0(names(m_list)[i], "s"),
			set_order = NULL, comb_order = NULL,
			top_annotation = HeatmapAnnotation(
				"Relative\nfraction" = anno_barplot(
					rel_comb_size[, i],
					ylim = c(0, 0.5),
					gp = gpar(fill = "black"),
					border = FALSE,
					height = unit(2, "cm"),
				), 
				annotation_name_side = "left",
				annotation_name_rot = 0),
			right_annotation = upset_right_annotation(m_list[[i]], 
				ylim = c(0, max_set_size))
		)
}
grid.newpage()
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
draw(ht_list1, newpage = FALSE)
popViewport()
pushViewport(viewport(x = 0.5, width = 0.5, just = "left"))
draw(ht_list2, newpage = FALSE)
popViewport()
```

Finally we can add the statistics of years, ratings and number of watches for
each combination set as boxplot annotations to the right of the UpSet plot.

```{r, fig.width = 6, fig.height = 8}
m = make_comb_mat(movies[, genre], remove_complement_set = TRUE)
m = m[comb_degree(m) > 0]
comb_elements = lapply(comb_name(m), function(nm) extract_comb(m, nm))
years = lapply(comb_elements, function(ind) movies$ReleaseDate[ind])
rating = lapply(comb_elements, function(ind) movies$AvgRating[ind])
watches = lapply(comb_elements, function(ind) movies$Watches[ind])

UpSet(t(m)) + rowAnnotation(years = anno_boxplot(years),
	rating = anno_boxplot(rating),
	watches = anno_boxplot(watches))
```

We can see the movies with genre "Scifi + Children" were produced quite old
but the ratings are not bad. The movies with genre "Action + Children" have
the lowest ratings.

## Example with the genomic regions

The H3K4me3 ChIP-seq peaks from six [Roadmap](http://www.roadmapepigenomics.org/) samples are
visualized by UpSet plot. The six samples are:

- [ESC, E016](https://egg2.wustl.edu/roadmap/data/byFileType/peaks/consolidated/narrowPeak/E016-H3K4me3.narrowPeak.gz)
- [ES-derived, E004](https://egg2.wustl.edu/roadmap/data/byFileType/peaks/consolidated/narrowPeak/E004-H3K4me3.narrowPeak.gz)
- [ES-derived, E006](https://egg2.wustl.edu/roadmap/data/byFileType/peaks/consolidated/narrowPeak/E006-H3K4me3.narrowPeak.gz)
- [Brain, E071](https://egg2.wustl.edu/roadmap/data/byFileType/peaks/consolidated/narrowPeak/E071-H3K4me3.narrowPeak.gz)
- [Muscle, E100](https://egg2.wustl.edu/roadmap/data/byFileType/peaks/consolidated/narrowPeak/E100-H3K4me3.narrowPeak.gz)
- [Heart, E104](https://egg2.wustl.edu/roadmap/data/byFileType/peaks/consolidated/narrowPeak/E104-H3K4me3.narrowPeak.gz)

First read the files and convert to `GRanges` objects.

```{r}
file_list = c(
	"ESC" = "data/E016-H3K4me3.narrowPeak.gz",
	"ES-deriv1" = "data/E004-H3K4me3.narrowPeak.gz",
	"ES-deriv2" = "data/E006-H3K4me3.narrowPeak.gz",
	"Brain" = "data/E071-H3K4me3.narrowPeak.gz",
	"Muscle" = "data/E100-H3K4me3.narrowPeak.gz",
	"Heart" = "data/E104-H3K4me3.narrowPeak.gz"
)
library(GenomicRanges)
peak_list = lapply(file_list, function(f) {
	df = read.table(f)
	GRanges(seqnames = df[, 1], ranges = IRanges(df[, 2], df [, 3]))
})
```

Make the combination matrix. Note now the size of the sets and the combination
sets are **total base pairs or the sum of width of the regions**. We only keep
the combination sets with more than 500kb.

```{r, fig.width = 8, fig.height = 4}
m = make_comb_mat(peak_list)
m = m[comb_size(m) > 500000]
UpSet(m)
```

We can nicely format the axis labels by setting `axis_param`:

```{r, fig.width = 8, fig.height = 4}
UpSet(m, 
	top_annotation = upset_top_annotation(
		m,
		axis_param = list(at = c(0, 1e7, 2e7),
			labels = c("0MB", "10MB", "20MB")),
		height = unit(4, "cm")
	),
	right_annotation = upset_right_annotation(
		m,
		axis_param = list(at = c(0, 2e7, 4e7, 6e7),
			labels = c("0MB", "20MB", "40MB", "60MB"),
			labels_rot = 0),
		width = unit(4, "cm")
	))
```

With each set of genomic regions, we can associate more information to it,
such as the mean methylation or the distance to nearest TSS.

```{r, fig.width = 10, fig.height = 4.2}
subgroup = c("ESC" = "group1",
	"ES-deriv1" = "group1",
	"ES-deriv2" = "group1",
	"Brain" = "group2",
	"Muscle" = "group2",
	"Heart" = "group2"
)
comb_sets = lapply(comb_name(m), function(nm) extract_comb(m, nm))
comb_sets = lapply(comb_sets, function(gr) {
	# we just randomly generate dist_to_tss and mean_meth
	gr$dist_to_tss = abs(rnorm(length(gr), mean = runif(1, min = 500, max = 2000), sd = 1000))
	gr$mean_meth = abs(rnorm(length(gr), mean = 0.1, sd = 0.1))
	gr
})
UpSet(m, 
	top_annotation = upset_top_annotation(
		m,
		axis_param = list(at = c(0, 1e7, 2e7),
			labels = c("0MB", "10MB", "20MB")),
		height = unit(4, "cm")
	),
	right_annotation = upset_right_annotation(
		m,
		axis_param = list(at = c(0, 2e7, 4e7, 6e7),
			labels = c("0MB", "20MB", "40MB", "60MB"),
			labels_rot = 0),
		width = unit(4, "cm")
	),
	left_annotation = rowAnnotation(group = subgroup[set_name(m)], show_annotation_name = FALSE),
	bottom_annotation = HeatmapAnnotation(
		dist_to_tss = anno_boxplot(lapply(comb_sets, function(gr) gr$dist_to_tss), outline = FALSE),
		mean_meth = sapply(comb_sets, function(gr) mean(gr$mean_meth)),
		annotation_name_side = "left"
	)
)
```
